---
title: "GMERT implementation"
author: "Paolo Colussi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
rm(list = ls())
library(rpart)   # CART trees used for the fixed (nonlinear) component f(X)
library(MASS)    # mvrnorm() to simulate correlated random effects for clusters
library(lme4)    # glmer() to fit standard GLMMs for comparison
library(parallel) # mclapply() for parallel computation
library(ggplot2) # plotting
```

## GMERT

Let $y_{ij}$ denote the binary response for the $j$-th observation ($j = 1, \ldots, n_i$) within cluster $i$ ($i = 1, \ldots, n$). 
Conditional on the random effects $b_i$, the responses $y_{ij}$ are assumed independent and follow a Bernoulli distribution with parameter $\mu_{ij}$, that is,
\[
y_{ij} \mid b_i \sim \text{Bernoulli}(\mu_{ij}), 
\qquad 
\mathbb{E}(y_{ij} \mid b_i) = \mu_{ij}, 
\qquad 
\text{Var}(y_{ij} \mid b_i) = \mu_{ij}(1 - \mu_{ij}).
\]

The relationship between the conditional mean $\mu_{ij}$ and the linear predictor $\eta_{ij}$ is specified through the canonical logit link function,
\[
\eta_{ij} = g(\mu_{ij}) = \log\!\left(\frac{\mu_{ij}}{1 - \mu_{ij}}\right).
\]

In the generalized mixed-effects regression tree (GMERT) model, the linear predictor is composed of a fixed component $f(x_{ij})$ and a random component $z_{ij}^{\top} b_i$, such that
\[
\eta_{ij} = f(x_{ij}) + z_{ij}^{\top} b_i,
\qquad 
b_i \sim \mathcal{N}(0, D),
\]
where $f(x_{ij})$ is an unknown function representing the fixed effects, estimated nonparametrically by means of a regression tree, and $b_i$ i

### Algorithm

The GMERT algorithm is the ML-based PQL-EM-algorithm in which we replace the linear structure used to estimate the fixed part of the model by a standard tree structure. The algorithm is as follows:

**Step 0**

Let $M=0$. Starting from initial values for the conditional means $\hat\mu_{ij}^{(0)}$ (for binary data, one convenient choice is $\hat\mu_{ij}^{(0)}=0.25$ if $y_{ij}=0$ and $\hat\mu_{ij}^{(0)}=0.75$ if $y_{ij}=1$), define the linearized (working) responses and the associated weights according to the penalized quasi-likelihood (PQL) scheme. Specifically, set
\[
\tilde y_i^{(0)} \;=\; g\!\big(\hat\mu_i^{(0)}\big) \;+\; \big(y_i-\hat\mu_i^{(0)}\big)\odot g'\!\big(\hat\mu_i^{(0)}\big),
\qquad 
W_i^{(0)} \;=\; \mathrm{diag}\!\big(w_{ij}^{(0)}\big),
\]
with $g(\cdot)$ the logit link, $g'(\mu_{ij}) = [\mu_{ij}(1-\mu_{ij})]^{-1}$, and $w_{ij}^{(0)} = \big(v_{ij}\,g'(\hat\mu_{ij}^{(0)})^2\big)^{-1}$. For Bernoulli outcomes, $v_{ij}=\mu_{ij}(1-\mu_{ij})$, which yields the familiar working weights $w_{ij}^{(0)}=\hat\mu_{ij}^{(0)}\!\big(1-\hat\mu_{ij}^{(0)}\big)$. Using $(\tilde y_i^{(0)}, W_i^{(0)})$, fit a \emph{weighted} linear mixed-effects pseudo-model to obtain initial estimates $\hat\sigma^{2(0)}$ and $\hat D^{(0)}$. These values initialize the subsequent weighted MERT iterations that update the tree-based fixed part and the random effects under the PQL framework.\footnote{See GMERT algorithm initialization and binary special case for the definitions of $\tilde y$ and $w_{ij}$.}

**Outer loop**

After obtaining the initial estimates of the working responses $\tilde y_i^{(0)}$, the weights $W_i^{(0)}$, and the variance components $\hat\sigma^{2(0)}$ and $\hat D^{(0)}$, the outer loop is initialized by setting $M = 0$. This loop governs the successive updates of the working quantities $(\eta_i, \mu_i, \tilde y_i, W_i)$ within the PQL framework. At each outer iteration, the algorithm calls the inner weighted MERT procedure to update the fixed and random components, after which the working responses and weights are recomputed. The outer loop proceeds until convergence of the linear predictors $\eta_i$.

**Start Inner loop**

While the outer loop governs the PQL updates of the working responses and weights, the inner loop performs the weighted mixed-effects regression tree estimation using the current pseudo-data. For each outer iteration $M$, and while the generalized log-likelihood (GLL) has not yet converged, the following steps are repeated.

**Step 1**

At the beginning of the $m$-th inner iteration, compute the adjusted working responses
\[
\tilde y_i^{\ast(m)} = \tilde y_i^{(M)} - Z_i\,\hat b_i^{(m-1)}, \qquad i = 1, \ldots, n.
\]
A weighted regression tree is then fitted to the data $(\tilde y_i^{\ast(m)}, X_i, W_i^{(M)})$ using a standard tree algorithm, thereby yielding an updated estimate of the fixed component, denoted $\hat f^{(m)}(X_i)$. 
The random effects are subsequently updated by the empirical best linear unbiased predictor (BLUP),
\[
\hat b_i^{(m)} = \hat D^{(m-1)}(W_i^{(M)1/2} Z_i)^{\top}
\hat V_i^{-(m-1)} \!\left( W_i^{(M)1/2}\tilde y_i^{(M)} - W_i^{(M)1/2}\hat f^{(m)}(X_i) \right),
\]

In R:
```{r b_fun}
# b_fun: updates random effects b_i given current D and provided V_i
#        w is a LIST with one diagonal weight matrix W_i per cluster
b_fun <- function(G,    # number of clusters
                  Z,    # N x q random-effects design (stacked Z_i)
                  Vi,   # list length G with V_i matrices (n_i x n_i) 
                  D,    # q x q covariance of random effects
                  idx,  # list length G with row indices per cluster
                  y_t,  # length N pseudo-response (y_tilde)
                  fhat, # length N fixed-part prediction
                  w     # length N weights 
){
  b <- matrix(0, G, ncol(Z))
  for (g in seq_len(G)) {
    Zi   <- Z[idx[[g]], , drop = FALSE]
    Wi_sq <- diag(sqrt(w[idx[[g]]]))                              # W_i^{1/2}

    Zi_w <- Wi_sq %*% Zi                                          # W^{1/2} Z_i
    rhs  <-  Wi_sq %*% y_t[idx[[g]]] - Wi_sq %*% fhat[idx[[g]]]   # W^{1/2} (y - fhat)

    b[g, ] <- D %*% t(Zi_w) %*% solve(Vi[[g]]) %*% rhs             # b_i = D (Z_w)^T V_i^{-1} rhs
  }
  b
}

# b_fun_small: optimized GMERT random-effects update (small q×q system)
b_fun_small <- function(G,        # number of clusters
                        Z,        # N x q random-effects design matrix
                        w,        # length N weights 
                        D,        # q x q covariance of random effects
                        sigma2,   # residual variance
                        idx,      # list of length G, indices per cluster
                        y_t,      # pseudo-response vector
                        fhat      # fitted fixed-part predictions
){
  b <- matrix(0, G, ncol(Z))      # store b_i estimates (G x q)

  for (g in seq_len(G)) {
    # --- extract cluster-specific data ---
    Zi <- Z[idx[[g]], , drop = FALSE]      # cluster design (n_i x q)
    y_res <- y_t[idx[[g]]] - fhat[idx[[g]]]# residuals (y_t - fhat)

    # --- compute small q×q system components ---
    # A_i = D^{-1} + (1/σ²) * Z_i^T W_i Z_i
    # rhs = (1/σ²) * Z_i^T W_i (y_t - fhat)
    A   <- solve(D) + (1 / sigma2) * crossprod(Zi, w[idx[[g]]] * Zi)
    rhs <- (1 / sigma2) * crossprod(Zi, w[idx[[g]]] * y_res)

    # --- solve for b_i via Cholesky (stable) ---
    R <- chol(A)
    b[g, ] <- backsolve(R, forwardsolve(t(R), rhs))
  }
  b
}

system.time(b_fun(G = G, Z = Z, Vi = Vi, D = D, idx = idx_by_cluster, y_t = y_t, fhat = fhat, w = w))
system.time(b_fun_small(G = G, Z = Z, D = D, idx = idx_by_cluster, y_t = y_t, fhat = fhat, w = w, sigma2 = sigma2))


```


where
\[
\hat V_i^{(m-1)} = W_i^{(M)1/2} Z_i\,\hat D^{(m-1)} (W_i^{(M)1/2} Z_i)^{\top} + \hat\sigma^{2(m-1)} I_{n_i}.
\]

In R:
```{r Vi_fun}
# Vi_fun: builds list of V_i = Z_i D Z_i^T + sigma^2 I_{n_i}
Vi_fun <- function(Z,   # Z   : N x q random-effects design
                   D,   # D   : q x q covariance of random effects
                   s2,  # s2  : scalar sigma^2 (residual variance; homoskedastic)
                   G,   # G   : number of clusters
                   idx, # idx : list of row indices per cluster
                   w    # length N weights 
                   ){
  Vi <- list()
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]                                    # Z_i
    Wi_sq <- diag(sqrt(w[idx[[g]]]))                                     # W_i^{1/2}
    Zi_w <- Wi_sq %*% Zi                                                 # W^{1/2} Z_i
    Vi[[g]] <- Zi_w %*% D %*% t(Zi_w) + as.numeric(s2) * diag(nrow(Zi_w))  # V_i
  }
  # Output:
  # list of length G with n_i x n_i covariance matrices V_i
  return(Vi)
}
```

**Step 2**

Given the new estimates of the fixed and random parts, the variance parameters are updated according to
\[
\hat\sigma^{2(m)} = \frac{1}{N}\sum_{i=1}^{n}
\left\{
\hat\varepsilon_i^{(m)\top}\hat\varepsilon_i^{(m)} + \hat\sigma^{2(m-1)}\!\left[n_i - \hat\sigma^{2(m-1)} \operatorname{tr}\!\big(\hat V_i^{-(m-1)}\big)\right]
\right\},
\]

with residuals
\[
\hat\varepsilon_i^{(m)} = W_i^{(M)1/2}\tilde y_i^{(M)} - W_i^{(M)1/2}\hat f^{(m)}(X_i) - W_i^{(M)1/2}Z_i\hat b_i^{(m)}.
\]


In R:
```{r sigma_fun}
# sigma_fun: updates sigma^2 using paper's closed-form expression
sigma_fun <- function(N,    # N   : total number of observations
                      G,    # G   : number of clusters
                      idx,  # idx : list of row indices per cluster
                      b,    # b   : matrix q × n vector of random effects 
                      y,    # y   : response vector length N
                      Z,    # Z   : N x q random-effects design
                      D,    # D   : q x q covariance of random effects
                      Vi,   # Vi  : list of length G with V_i matrices (n_i x n_i)
                      s2,   # s2  : scalar sigma^2 (residual variance; homoskedastic)
                      fhat, # fhat: current fixed-effect fit
                      w     # length N weights 
                      ){
  term_sum <- 0
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]
    Wi_sq <- diag(sqrt(w[idx[[g]]]))                    # W_i^{1/2}                         
    e <- Wi_sq %*% y[idx[[g]]] - Wi_sq %*% fhat[idx[[g]]] - Wi_sq %*% Zi%*% b[g, ]       #  weighted residuals epsilon_i = W_i^{1/2} y - W_i^{1/2} fhat - W_i^{1/2} Z_i b_i
    term_sum <- term_sum + crossprod(e) + s2 * (length(idx[[g]]) -
                       s2 * sum(diag(solve(Vi[[g]]))))     # add cluster-i contribution
  }
  # Output:
  #   scalar sigma^2 new estimate
  term_sum / N                                             # divide by total N
}
```

\[
\hat D^{(m)} = \frac{1}{n}\sum_{i=1}^{n}
\left\{
\hat b_i^{(m)}\hat b_i^{(m)\top}
+ \Big[\hat D^{(m-1)} - \hat D^{(m-1)} (W_i^{(M)1/2} Z_i)^{\top}
\hat V_i^{-(m-1)} W_i^{(M)1/2} Z_i \hat D^{(m-1)}\Big]
\right\},
\]

In R:
```{r D_fun}
# D_fun: updates D using the paper's expression (average of b_i b_i^T plus shrinkage term)
D_fun <- function(G,   # G   : number of clusters
                  idx, # idx : list of row indices per cluster
                  b,   # b   : matrix q × n vector of random effects 
                  Z,   # Z   : N x q random-effects design
                  D,   # D   : q x q covariance of random effects
                  Vi,  # Vi  : list of length G with V_i matrices (n_i x n_i)
                  w    # length N weights 
                  ){
  q <- ncol(Z)
  S <- matrix(0, q, q)                                     # accumulator
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]                      # Z_i
    Wi_sq <- diag(sqrt(w[idx[[g]]]))                       # W_i^{1/2}
    Zi_w <- Wi_sq %*% Zi                                   # W^{1/2} Z_i
    S <- S + tcrossprod(b[g, ]) +                          # b_i b_i^T
      (D - D %*% t(Zi_w) %*% solve(Vi[[g]]) %*% Zi_w %*% D)    # shrinkage term per cluster
  }
  # Output:
  #   q x q updated D
  S/G                                                      # average over clusters
}
```

**Step 3**

At each inner iteration, the following generalized log-likelihood (GLL) is computed:
\[
\mathrm{GLL}\big(f(X_i), b_i \mid y\big) =
\sum_{i=1}^{n}
\left\{
\hat\varepsilon_i^{(m)\top}(\hat\sigma^{2(m)}I_{n_i})^{-1}\hat\varepsilon_i^{(m)}
+ \hat b_i^{(m)\top}\hat D^{-(m)}\hat b_i^{(m)}
+ \log|\hat D^{(m)}| + \log|\hat\sigma^{2(m)} I_{n_i}|
\right\}.
\]


In R:
```{r gll_fun}

# gll_fun: computes generalized log-likelihood (up to additive constant)
gll_fun <- function(idx, # idx : list of row indices per cluster
                    b,   # b   : matrix q × n vector of random effects 
                    y,   # y   : response vector length N
                    Z,   # Z   : N x q random-effects design
                    D,   # D   : q x q covariance of random effects
                    s2,  # s2  : scalar sigma^2 (residual variance; homoskedastic)
                    fhat,# fhat: current fixed-effect fit
                    w    # length N weights 
                    ){
  logdetD <- as.numeric(determinant(D, logarithm = TRUE)$modulus)     # log|D|
  term_b <- sum(rowSums((b %*% solve(D)) * b))                         # sum_i b_i^T D^{-1} b_i
  term_r <- 0
  term_logRi <- 0
  for (i in seq_along(idx)) {
    Zi <- Z[idx[[i]], , drop = FALSE]                      # Z_i
    Wi_sq <- diag(sqrt(w[idx[[i]]]))                       # W_i^{1/2}                         
    e <- Wi_sq %*% y[idx[[i]]] - Wi_sq %*% fhat[idx[[i]]] - Wi_sq %*% Zi%*% b[i, ]       #  weighted residuals epsilon_i = W_i^{1/2} y - W_i^{1/2} fhat - W_i^{1/2} Z_i b_i
    term_r <- term_r + sum(e^2) / s2                                  # (1/s2) * ||ri||^2  
    term_logRi <- term_logRi + length(idx[[i]]) * log(s2)              # n_i * log(s2)
  }
  
  # Output:
  #   scalar GLL value (up to constant)
  # GLL = sum_i [ (y - f - Zb)^T R_i^{-1} (y - f - Zb) + log|R_i| ] +
  # + sum_i b_i^T D^{-1} b_i + G*log|D|
  term_r + term_logRi + term_b + length(idx) * logdetD                 # assemble GLL
}
```

**End Inner loop**

The inner loop continues until the relative change in the GLL between two successive iterations is smaller than a predefined tolerance threshold. To check convergence we use the Aitken stop, *this is not in the paper*, and I'm really unsure if it good enough, but looks like it (*Indeed for now it's not used anymore, still here because why not*)

In R:

```{r ait_stop}
# ait_stop: Aitken acceleration based stopping rule
ait_stop <- function(gll,  #gll : numeric vector with GLL history (needs last three values)
                     tol,  #tol : tolerance for relative error to asymptote
                     it    #it  : current iteration index 
                     ){
  converged <- FALSE
  l1 <- gll[it + 2] - gll[it + 1]                                      # L_t
  l2 <- gll[it + 1] - gll[it]                                          # L_{t-1}
  a <- l1 / l2                                                         # Aitken ratio
  if (is.finite(a) && a > 0 && a < 1) {                                # only meaningful if 0<a<1
    L_inf <- gll[it + 1] + l1 / (1 - a)                                # extrapolate asymptote
    err <- abs(L_inf - gll[it + 2]) / (abs(L_inf) + 1e-12)             # relative error to asymptote
    if (err < tol) converged <- TRUE                                   # declare convergence
  }
  converged                                                            # return flag
}
```

**End Outer loop**

After convergence of the inner loop at outer iteration $M$, the working quantities are updated to prepare the next outer iteration. 
The linear predictors are first recomputed as
\[
\hat\eta_i^{(M+1)} = \hat f^{(m)}(X_i) + Z_i\,\hat b_i^{(m)}, \qquad i = 1, \ldots, n,
\]
and the conditional means are obtained by applying the inverse of the logit link,
\[
\hat\mu_i^{(M+1)} = g^{-1}\!\big(\hat\eta_i^{(M+1)}\big) = 
\frac{\exp\!\big(\hat\eta_i^{(M+1)}\big)}{1 + \exp\!\big(\hat\eta_i^{(M+1)}\big)}.
\]
The new pseudo-responses and weights are then derived as
\[
\tilde y_i^{(M+1)} = g\!\big(\hat\mu_i^{(M+1)}\big) 
+ \big(y_i - \hat\mu_i^{(M+1)}\big)\, g'\!\big(\hat\mu_i^{(M+1)}\big),
\qquad 
W_i^{(M+1)} = \mathrm{diag}\!\big(w_{ij}^{(M+1)}\big),
\]
with $w_{ij}^{(M+1)} = \hat\mu_{ij}^{(M+1)} \big(1 - \hat\mu_{ij}^{(M+1)}\big)$. 
The index $M$ is then increased by one, and the algorithm returns to the inner loop until the convergence of the linear predictors $\eta_i$ across successive outer iterations.

### Implementation

Here we put the algorithm in a single function with many parameter to be able to control the convergence and to be able to tune the tree from step 1. Also we have a convergence plot for the GLL, if we want to check visually the convergence. 

```{r fit_fun}
fit_gmert    <- function(df,               # df: data.frame with columns
                                           #   id   : cluster identifier (factor or integer)
                                           #   y    : numeric response
                                           #   x1,x2,x3 : numeric covariates
                                           #   leaf : true generating leaf (optional, for diagnostics)
                         max_iter_inn = 1000,  # maximum number of EM iterations (inner loop)
                         max_iter_out = 1000,  # maximum number of PQL iterations (outer loop)
                         tol = 1e-6,           # convergence tolerance for both loops (Aitken or relative diff)
                         cp = 0.0,             # rpart complexity parameter (pruning threshold)
                         minsplit = 50,        # minimum number of obs required to attempt a split
                         minbucket = 20,       # minimum number of obs in any terminal node
                         maxdepth = 5,         # maximum tree depth
                         xval = 10             # number of cross-validation folds in rpart
                         ) {

  # --- Basic setup ---
  N <- nrow(df)                                 # total number of observations
  G <- length(unique(df$id))                    # number of clusters
  idx_by_cluster <- split(seq_len(N), df$id)    # list: row indices grouped by cluster

  y <- df$y                                     # response vector
  Z <- cbind(1, df$x1)                          # random-effects design: intercept + slope on x1
  q <- ncol(Z)                                  # number of random effects (q = 2)

  # --- Initialization (Step 0) ---
  M <- 0L                                       # outer-loop counter
  mu <- ifelse(y == 1, 0.75, 0.25)              # initial conditional means
  y_t <- log(mu / (1 - mu)) + (y - mu) / (mu * (1 - mu))  # initial pseudo-response (PQL linearization)
  w <- mu * (1 - mu)                            # initial working weights
  sigma2 <- 1                                   # initial residual variance
  D <- diag(2)                                  # initial random-effects covariance (identity)
  b <- matrix(0, G, q)                          # initialize cluster random effects
  gll <- c(0, 0)                                # GLL storage (2 slots for Aitken acceleration)
  eta_old <- rep(0, N)                          # previous eta for outer-loop convergence check
  converged_in <- c()                           # inner-loop convergence flags (per outer iteration)
  converged_out <- FALSE                        # outer-loop convergence flag

  # --- Outer loop (PQL updates) ---
  repeat {
    m = 0                                       # reset inner-loop counter

    # --- Inner loop (EM-like iterations) ---
    repeat {
      m <- m + 1L

      # (1.i) Partial E-step: compute adjusted pseudo-response y_tilde* = y_tilde - Z b
      zb <- numeric(N)                          # cluster-specific random contributions
      for (g in seq_len(G)) {
        idx <- idx_by_cluster[[g]]              # indices for cluster g
        zb[idx] <- Z[idx, , drop = FALSE] %*% b[g, ]  # Z_i b_i
      }
      y_star <- y_t - zb                        # adjusted response for tree fit

      # (1.ii) M-step: fit regression tree for fixed effects f(X)
      ctrl <- rpart.control(cp = cp, minsplit = minsplit, xval = xval,
                            minbucket = minbucket, maxdepth = maxdepth)
      Xdf <- data.frame(x1 = df$x1, x2 = df$x2, x3 = df$x3)
      tree <- rpart(y_star ~ x1 + x2 + x3,
                    data = cbind(y_star = y_star, Xdf),
                    weights = w, method = "anova", control = ctrl)
      fhat <- as.numeric(predict(tree, newdata = Xdf))

      # (1.iii) Update random effects b_i
      Vi <- Vi_fun(Z = Z, D = D, s2 = sigma2, G = G, idx = idx_by_cluster, w = w)  # build V_i per cluster
      b <- b_fun(G = G, Z = Z, Vi = Vi, D = D, idx = idx_by_cluster,
                 y_t = y_t, fhat = fhat, w = w)                                   # update b_i estimates

      # (2.i) Update sigma^2 (residual variance)
      sigma2 <- sigma_fun(N = N, G = G, idx = idx_by_cluster,
                          b = b, y = y_t, Z = Z, D = D,
                          Vi = Vi, s2 = sigma2, fhat = fhat, w = w)

      # (2.ii) Update D (random-effects covariance)
      D <- D_fun(G = G, idx = idx_by_cluster,
                 Z = Z, D = D, Vi = Vi, b = b, w = w)

      # --- Inner-loop convergence check (GLL stabilization) ---
      gll[m + 2] <- gll_fun(D = D, b = b, idx = idx_by_cluster,
                            Z = Z, y = y_t, fhat = fhat, s2 = sigma2, w = w)
      if (m > 1L) {
        rel <- abs(gll[m + 2] - gll[m + 1]) / (abs(gll[m + 1]) + 1e-12)
        if (rel < tol) { n_iter <- m; converged_in_t <- TRUE; break }
      }

      if (m >= max_iter_inn) {                  # max-iteration guard
        n_iter <- m
        converged_in_t <- FALSE
        # message(sprintf("WARNING: the EM algorithm did not converge in %d iterations.", max_iter_inn))
        break
      }
    }

    # --- Outer-loop update (PQL step) ---
    converged_in <- c(converged_in, converged_in_t)
    zb <- numeric(N)
    for (g in seq_len(G)) {
      idx <- idx_by_cluster[[g]]
      zb[idx] <- Z[idx, , drop = FALSE] %*% b[g, ]
    }
    eta <- fhat + zb                            # recompute linear predictor

    # (Outer stopping rule – paper style)
    d_eta <- sqrt(mean((eta - eta_old)^2))      # RMS change of eta
    if (d_eta < tol) {
      converged_out <- TRUE
      break
    }

    # Update working quantities for next outer iteration
    eta_old <- eta
    mu <- exp(eta) / (1 + exp(eta))             # updated conditional means
    y_t <- log(mu / (1 - mu)) + (y - mu) / (mu * (1 - mu))  # new pseudo-response
    w <- mu * (1 - mu)                          # new working weights

    M <- M + 1L
    if (M >= max_iter_out) {                    # guard against outer non-convergence
      converged_out <- FALSE
      message(sprintf("WARNING: the PQL algorithm did not converge in %d iterations.", max_iter_out))
      break
    }
  }

  # --- Return fitted components ---
  list(
    tree = tree,                   # fitted rpart tree (fixed-effects function f(X))
    b = b,                         # estimated random effects (G x q)
    D = D,                         # estimated random-effects covariance
    sigma2 = sigma2,               # estimated residual variance
    mu = mu,                       # conditional means
    converged_in = converged_in,   # convergence flags for inner loops
    converged_out = converged_out, # convergence flag for outer loop
    n_iter = n_iter,               # iterations performed (inner loop)
    train_ids = df$id,             # cluster identifiers in training set
    gll = gll,                     # GLL trace (for diagnostics)
    tol = tol                      # convergence tolerance used
  )
}

```

### Prediction

To predict the response for a new observation that belongs to a cluster among those used to fit the MERT model, we use both its corresponding population-averaged tree prediction and the predicted random part corresponding to its cluster. For a new observation that belongs to a cluster not included in the sample used to estimate the model parameters, we can only take the corresponding population-averaged tree prediction.

```{r}
predict_gmert <- function(fit, new_df, thr = 0.5  
) {
  # fixed part: regression tree predictions using predictors only
  fhat <- as.numeric(predict(fit$tree, new_df))        # tree fitted on y*; use on new data
  # construct random effects design for new data: [1, x1]
  Znew <- cbind(1, new_df$x1)
  # container for random effects contributions
  add <- numeric(nrow(new_df))
  # clusters seen during training (so we have estimated b_i)
  clus_fit <- sort(unique(fit$train_ids))
  # map each new cluster id to its index in clus_fit (NA if unseen)
  map <- match(new_df$id, clus_fit)
  seen <- !is.na(map)                                  # TRUE for rows belonging to seen clusters
  if (any(seen)) {
    # add random effect contribution: row-wise sum of [1, x1] * b_i
    add[seen] <- rowSums(Znew[seen, , drop = FALSE] *
                         fit$b[map[seen], ])
  }
  # total prediction = fixed part + random effects contribution
  eta   <- fhat + add
  p     <- 1 / (1 + exp(-eta))

  # return 0/1; 
  yhat  <- ifelse(p >= thr, 1, 0)
  yhat
}
```


## Simulation

In this section, we generate a clustered binary dataset designed to mimic the structure assumed by the GMERT model.  
Each cluster has its own random intercept and random slope for the covariate `x1`.  
The binary response `y` is simulated under a logistic link function that combines fixed effects and random effects.  
This setup provides a realistic test bed for evaluating mixed-effects tree algorithms such as GMERT and standard GLMMs.



```{r}
# ---------------------------------------------------------------
# sim_data_gmert: simulate clustered binary data for GMERT models
# ---------------------------------------------------------------
sim_data_gmert <- function(G = 50,          # number of clusters
                           n_i = 40,        # observations per cluster
                           beta0 = 0.5,     # fixed intercept
                           beta1 = 1.2,     # fixed effect for x1
                           beta2 = -0.8,    # fixed effect for x2
                           beta3 = 0.6,     # fixed effect for x3
                           sigma_b0 = 0.8,  # SD of random intercept
                           sigma_b1 = 0.5,  # SD of random slope for x1
                           rho = 0.2,       # correlation between b0 and b1
                           seed = 123) {    # random seed for reproducibility

  set.seed(seed)

  # covariance matrix of random effects (b0_i, b1_i)
  D <- matrix(c(sigma_b0^2, rho * sigma_b0 * sigma_b1,
                rho * sigma_b0 * sigma_b1, sigma_b1^2), 2, 2)

  # generate random effects for G clusters
  b <- MASS::mvrnorm(G, mu = c(0, 0), Sigma = D)

  # cluster identifiers
  id <- rep(1:G, each = n_i)

  # generate covariates: x1 (uniform), x2 (normal), x3 (binary)
  x1 <- runif(G * n_i, -2, 2)
  x2 <- rnorm(G * n_i, 0, 1)
  x3 <- rbinom(G * n_i, 1, 0.5)

  # linear predictor: fixed part + random part (b0_i + b1_i * x1)
  eta <- numeric(G * n_i)
  for (g in seq_len(G)) {
    idx <- which(id == g)
    eta[idx] <- beta0 + beta1 * x1[idx] + beta2 * x2[idx] + beta3 * x3[idx] +
                b[g, 1] + b[g, 2] * x1[idx]
  }

  # convert to probabilities via logistic link
  p <- 1 / (1 + exp(-eta))

  # binary outcome drawn from Bernoulli(p)
  y <- rbinom(G * n_i, 1, p)

  # return as data.frame ready for model fitting
  data.frame(
    id = factor(id),
    y = y,
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
}

```

## Example run

This section shows a complete example of fitting the GMERT algorithm on a single simulated dataset.  
We start by generating clustered binary data that follow the same structure assumed by the model (binary outcome, logit link, and random intercept and slope on `x1`).  
We then fit the model with `fit_mert_RIC()` and inspect convergence diagnostics, estimated parameters, and predicted classifications.  
No train/test split is used here: the goal is to illustrate the fitting procedure itself and verify that the algorithm converges correctly.

```{r}
# ---------------------------------------------------------------
# EXAMPLE RUN
# ---------------------------------------------------------------

#  Simulate clustered binary data (30 clusters, 60 observations each)
df <- sim_data_gmert(G = 30, n_i = 60)

#  Fit the GMERT model using the full dataset
fit <- fit_gmert(df, max_iter_inn = 1000, 
                 max_iter_out = 10, tol = 1e-6)

#  Check convergence and model components
fit$converged_out       # whether the outer (PQL) loop converged
mean(fit$converged_in)  # proportion of inner (EM) loops that converged
head(fit$b)             # first rows of random-effect estimates (intercept, slope on x1)
fit$D                   # estimated covariance matrix of random effects
fit$sigma2              # estimated residual variance

```

`fit$b` shows the cluster-specific random intercepts and slopes, while `fit$D` and `fit$sigma2` summarize the estimated variability in the random structure.

```{r}
#    GLMM baseline with the same random-effects structure
#    (logistic link; random intercept + random slope on x1 by cluster id)
glmm_fit <- glmer(
  y ~ x1 + x2 + x3 + (1 + x1 | id),
  data = df,
  family = binomial(link = "logit")
)

summary(glmm_fit)  # fixed effects, random effects, VarCorr, etc.

```

The GLMM provides a parametric baseline (linear fixed effects) against which to compare GMERT’s nonparametric fixed part.  
Both models share the same random-effects structure, so differences in performance are attributable to how the fixed effects are modeled.

```{r}
# Predictions (binary) for both models on the same data
# GMERT: predict_mert_RIC returns 0/1 (no unseen clusters here)
y_hat_gmert <- predict_gmert(fit, df)

# GLMM: predict probabilities then threshold to 0/1
p_glmm      <- as.numeric(predict(glmm_fit, type = "response"))
y_hat_glmm  <- as.integer(p_glmm >= 0.5)

# Confusion matrices and metrics

# f1 score function
f1_fun <- function(table){
  precision <- table[2,2] / (sum(table[2, ]))
  recall    <- table[2,2] / (sum(table[, 2]))
  F1        <- 2 * precision * recall / (precision + recall)
  F1
}

# GMERT
cm_gmert <- table(predicted = y_hat_gmert, actual = df$y)
acc_gmert <- mean(y_hat_gmert == df$y)
f1_gmert  <- f1_fun(cm_gmert)

# confusion matrix for GMERT
cm_gmert

# accuracy and F1 for GMERT
acc_gmert
f1_gmert

# GLMM
cm_glmm <- table(predicted = y_hat_glmm, actual = df$y)
acc_glmm <- mean(y_hat_glmm == df$y)
f1_glmm  <- f1_fun(cm_glmm)

# confusion matrix for GLMM
cm_glmm

# accuracy and F1 for GLMM
acc_glmm
f1_glmm

```

**Confusion matrices** summarize the counts of TP/FP/FN/TN for each model.  
We report **accuracy** (overall correctness) and **F1**, the harmonic mean of precision and recall, which is more informative when classes are imbalanced.  




## Train/Test split

Here we repeat the analysis with a cluster-balanced train/test setup.  
Each cluster contributes the same proportion of observations to training and test, so the random effects `b_i` are estimable from training for **every** cluster that appears in test.  
We fit GMERT on the training set, fit a GLMM baseline, predict on the test set, and summarize performance with confusion matrices, accuracy, and F1.

```{r}
# ---------------------------------------------------------------
# Train/test split per cluster using the sim_data_gmert() output
# ---------------------------------------------------------------
split_gmert_data <- function(df, train_prop = 0.7, seed = 123) {
  set.seed(seed)

  train_idx <- integer(0)
  test_idx  <- integer(0)
  ids <- unique(df$id)

  for (g in ids) {
    idx_g <- which(df$id == g)
    n_train_g <- floor(length(idx_g) * train_prop)
    train_g <- sample(idx_g, n_train_g)
    test_g  <- setdiff(idx_g, train_g)
    train_idx <- c(train_idx, train_g)
    test_idx  <- c(test_idx, test_g)
  }

  train_df <- df[train_idx, ]
  test_df  <- df[test_idx, ]
  list(train = train_df, test = test_df)
}

```


```{r}
# ---------------------------------------------------------------
# TRAIN/TEST SPLIT EXAMPLE
# ---------------------------------------------------------------

# Simulate clustered binary data
set.seed(2)
data_full <- sim_data_gmert(G = 30, n_i = 60)

# Cluster-balanced split: fixed proportion per cluster in train
split_sets <- split_gmert_data(data_full, train_prop = 0.7, seed = 2)
data_train <- split_sets$train
data_test  <- split_sets$test

# Fit GMERT model on training data
model_gmert <- fit_gmert(data_train, max_iter_inn = 1000, 
                         max_iter_out = 10, tol = 1e-6)

# Check convergence and parameter estimates
model_gmert$converged_out        # outer-loop convergence
mean(model_gmert$converged_in)   # proportion of inner loops converged
model_gmert$D                    # random-effects covariance
model_gmert$sigma2               # residual variance

# Fit GLMM baseline with same random-effects structure
model_glmm <- glmer(
  y ~ x1 + x2 + x3 + (1 + x1 | id),
  data = data_train,
  family = binomial(link = "logit")
)
```

Both models are trained on the **same** training data. 

```{r}
# Predictions on the test data
# GMERT: binary predictions (0/1)
pred_gmert <- predict_gmert(model_gmert, data_test)

# GLMM: predict probabilities then threshold to 0/1
prob_glmm <- as.numeric(predict(model_glmm, newdata = data_test, type = "response"))
pred_glmm <- as.integer(prob_glmm >= 0.5)

# Confusion matrices and performance metrics

# GMERT metrics
cm_gmert <- table(predicted = pred_gmert, actual = data_test$y)
acc_gmert <- mean(pred_gmert == data_test$y)
f1_gmert  <- f1_fun(cm_gmert)

# confusion matrix for GMERT
cm_gmert

# accuracy and F1 for GMERT
acc_gmert
f1_gmert

# GLMM metrics
cm_glmm <- table(predicted = pred_glmm, actual = data_test$y)
acc_glmm <- mean(pred_glmm == data_test$y)
f1_glmm  <- f1_fun(cm_glmm)

# confusion matrix for GLMM
cm_glmm

# accuracy and F1 for GLMM
acc_glmm
f1_glmm

```


**Confusion matrices** show true/false positives and negatives on the test set.  
We report **accuracy** (overall correctness) and **F1**, which balances precision and recall.  
Since the split preserves cluster proportions across train and test, all test clusters are seen during training, allowing correct evaluation of cluster-specific random effects.

## Cross-validated predictions

## 10-Fold Cross-Validation

In this section, we evaluate the out-of-sample performance of GMERT against a standard GLMM baseline using a **10-fold cross-validation** procedure.

The goal is to assess how well each model generalizes to new observations drawn from the **same set of clusters**.  
To ensure that random effects remain estimable in every training fold, the data are split in a **cluster-balanced** way:  
each cluster contributes approximately one-tenth of its observations to every test fold, while the remaining data are used for training.  
This guarantees that all clusters appear in both the training and test partitions of every fold.

For each fold, we:
1. Fit the GMERT model on the training portion.
2. Fit a GLMM baseline with the same random-effects structure `(1 + x1 | id)`.
3. Predict binary outcomes on the test portion.
4. Compute the **confusion matrix**, **accuracy**, and **F1 score** for both models.

Finally, we aggregate the metrics across folds to compare the models in terms of mean and variability of their predictive performance.  
This setup provides a fair, cluster-consistent cross-validation framework to benchmark GMERT against the parametric GLMM.

```{r}
# ================================================================
# 10-fold cluster-balanced cross-validation: GMERT vs GLMM
# ================================================================

set.seed(42)  # CV reproducibility

# ----------------------------------------------------------------
# Simulate ONE dataset (all folds are made from the same df)
# ----------------------------------------------------------------
df_cv <- sim_data_gmert(G = 30, n_i = 60)  # clustered binary data

# ----------------------------------------------------------------
#   Build K cluster-aware folds
#    - Within each cluster g, randomly partition its indices into K slices
#    - Fold k is the union of the k-th slice from EVERY cluster
# ----------------------------------------------------------------
K <- 10
folds <- vector("list", K)                     # folds[[k]] will hold test row indices for fold k

for (g in unique(df_cv$id)) {
  idx_g <- which(df_cv$id == g)        # all rows belonging to cluster g
  idx_g <- sample(idx_g)               # shuffle rows within this cluster

  # cut() assigns approximately equal-sized groups labeled 1..K
  # (some clusters may not divide evenly; that's fine)
  slice_g <- split(idx_g, cut(seq_along(idx_g), K, labels = FALSE))

  # append cluster g's k-th slice to the global k-th fold
  for (k in seq_len(K)) {
    folds[[k]] <- c(folds[[k]], slice_g[[k]])
  }
}

cv_results <- vector("list", K)   # each element stores full fold results

# ----------------------------------------------------------------
#    Cross-validation loop
#    For k = 1..K:
#      - Train GMERT + GLMM on rows not in fold k
#      - Predict on rows in fold k
#      - Compute confusion matrices, accuracy, F1
# ----------------------------------------------------------------

for (k in seq_len(K)) {
  # ----- split rows for this fold
  test_idx  <- folds[[k]]
  train_idx <- setdiff(seq_len(nrow(df)), test_idx)

  df_train <- df[train_idx, ]
  df_test  <- df[test_idx, ]

  # ----- (a) GMERT: fit on training rows
  fit_g <- fit_gmert(df_train, max_iter_out = 10)

  # Predict on test rows (binary 0/1; all clusters exist in train by construction)
  yhat_g <- predict_gmert(fit_g, df_test)

  # ----- (b) GLMM: same random-effects structure, fit on training rows
  fit_l <- glmer(
    y ~ x1 + x2 + x3 + (1 + x1 | id),
    data = df_train,
    family = binomial(link = "logit"),
    control = glmerControl(optimizer = "bobyqa")
  )

  # Predict probabilities on test, then threshold at 0.5 → 0/1
  phat_l <- as.numeric(predict(fit_l, newdata = df_test, type = "response"))
  yhat_l <- as.integer(phat_l >= 0.5)

  # ----- (c) Metrics on the test portion of this fold
  # Confusion matrices
  cm_g <- table(predicted = yhat_g, actual = df_test$y)
  cm_l <- table(predicted = yhat_l, actual = df_test$y)

  # Accuracy
  acc_g <- mean(yhat_g == df_test$y)
  acc_l <- mean(yhat_l == df_test$y)

  # F1
  f1_g <- f1_fun(cm_g)
  f1_l <- f1_fun(cm_l)

  # ----- Store fold results
  cv_results[[k]] <- list(
    fold_id   = k,
    train_idx = train_idx,
    test_idx  = test_idx,
    gmert_fit = fit_g,
    glmm_fit  = fit_l,
    metrics = list(
      acc_gmert = acc_g, f1_gmert = f1_g,
      acc_glmm  = acc_l,  f1_glmm  = f1_l
    ),
    confusion = list(
      gmert = cm_g,
      glmm  = cm_l
    )
  )
}

# ----------------------------------------------------------------
#  Aggregate results across folds: mean ± sd for each model
# ----------------------------------------------------------------
# Extract metrics into a data.frame
metrics_df <- do.call(rbind, lapply(cv_results, function(x) {
  data.frame(
    fold = x$fold_id,
    acc_gmert = x$metrics$acc_gmert,
    f1_gmert  = x$metrics$f1_gmert,
    acc_glmm  = x$metrics$acc_glmm,
    f1_glmm   = x$metrics$f1_glmm
  )
}))

# Mean ± sd summary
cv_summary <- data.frame(
  model   = c("GMERT","GLMM"),
  acc_mean = c(mean(metrics_df$acc_gmert), mean(metrics_df$acc_glmm)),
  acc_sd   = c(sd(metrics_df$acc_gmert),   sd(metrics_df$acc_glmm)),
  f1_mean  = c(mean(metrics_df$f1_gmert),  mean(metrics_df$f1_glmm)),
  f1_sd    = c(sd(metrics_df$f1_gmert),    sd(metrics_df$f1_glmm))
)

# Inspect summary
cv_summary
metrics_df

```

Now, we do basically the same thing but in parallel

```{r, eval=FALSE}
# ================================================================
# 10-fold cluster-balanced CV in PARALLEL: GMERT vs GLMM
# ================================================================

set.seed(42)

# --- 0) Simulate one dataset ------------------------------------------------------
df <- sim_data_gmert(G = 30, n_i = 60)

# --- 1) Build K cluster-aware folds ----------------------------------------------
K <- 10
folds <- vector("list", K); for (k in seq_len(K)) folds[[k]] <- integer(0)

for (g in unique(df$id)) {
  idx_g <- which(df$id == g)
  idx_g <- sample(idx_g)                                # shuffle within cluster
  split_g <- split(idx_g, cut(seq_along(idx_g), K, labels = FALSE))
  for (k in seq_len(K)) folds[[k]] <- c(folds[[k]], split_g[[k]])
}

# ================================================================
# PARALLEL SECTION
# ================================================================

# pick number of workers (leave one core free when possible)
n_workers <- max(1L, min(K, detectCores() - 1L))
cl <- makeCluster(n_workers, type = "PSOCK")

# make RNG reproducible across workers
clusterSetRNGStream(cl, iseed = 42)

# load packages and export needed objects/functions to workers
clusterEvalQ(cl, { library(rpart); library(MASS);library(lme4) })
clusterExport(
  cl,
  varlist = c("df", "folds", "fit_gmert", "predict_gmert", "f1_fun", "K","Vi_fun",
                "b_fun","D_fun","sigma_fun","gll_fun"),
  envir = environment()
)

# --- Run folds in parallel -----------------------------------------------------
cv_results <- parLapply(cl, seq_len(K), function(k) {
  # define train/test indices
  test_idx  <- folds[[k]]
  train_idx <- setdiff(seq_len(nrow(df)), test_idx)
  df_train  <- df[train_idx, ]
  df_test   <- df[test_idx, ]

  # GMERT: fit and predict
  fit_gmert_k <- fit_gmert(df_train, max_iter_out = 2000)
  yhat_gmert  <- predict_gmert(fit_gmert_k, df_test)
  cm_gmert    <- table(predicted = yhat_gmert, actual = df_test$y)
  acc_gmert   <- mean(yhat_gmert == df_test$y)
  f1_gmert    <- f1_fun(cm_gmert)

  # GLMM: fit and predict (same random-effects structure)
  fit_glmm_k <- glmer(
    y ~ x1 + x2 + x3 + (1 + x1 | id),
    data = df_train,
    family = binomial(link = "logit"),
    control = glmerControl(optimizer = "bobyqa")
  )
  p_glmm     <- as.numeric(predict(fit_glmm_k, newdata = df_test, type = "response"))
  yhat_glmm  <- as.integer(p_glmm >= 0.5)
  cm_glmm    <- table(predicted = yhat_glmm,  actual = df_test$y)
  acc_glmm   <- mean(yhat_glmm == df_test$y)
  f1_glmm    <- f1_fun(cm_glmm)

  # return full fold bundle
  list(
    fold_id   = k,
    train_idx = train_idx,
    test_idx  = test_idx,
    gmert_fit = fit_gmert_k,
    glmm_fit  = fit_glmm_k,
    metrics = list(
      acc_gmert = acc_gmert, f1_gmert = f1_gmert,
      acc_glmm  = acc_glmm,  f1_glmm  = f1_glmm
    ),
    confusion = list(
      gmert = cm_gmert,
      glmm  = cm_glmm
    )
  )
})

# ---  Aggregate metrics across folds -------------------------------------------
metrics_df <- do.call(rbind, lapply(cv_results, function(x) {
  data.frame(
    fold      = x$fold_id,
    acc_gmert = x$metrics$acc_gmert,
    f1_gmert  = x$metrics$f1_gmert,
    acc_glmm  = x$metrics$acc_glmm,
    f1_glmm   = x$metrics$f1_glmm
  )
}))

cv_summary <- data.frame(
  model    = c("GMERT","GLMM"),
  acc_mean = c(mean(metrics_df$acc_gmert), mean(metrics_df$acc_glmm)),
  acc_sd   = c(sd(metrics_df$acc_gmert),   sd(metrics_df$acc_glmm)),
  f1_mean  = c(mean(metrics_df$f1_gmert),  mean(metrics_df$f1_glmm)),
  f1_sd    = c(sd(metrics_df$f1_gmert),    sd(metrics_df$f1_glmm))
)

```


```{r}
# ---------------------------------------------------------------
# Per-fold summary table from cv_results
# ---------------------------------------------------------------

fold_table <- do.call(rbind, lapply(cv_results, function(x) {
  # guard in case converged_in is missing/NULL
  inner_vec <- x$gmert_fit$converged_in
  inner_pct <- if (length(inner_vec)) mean(inner_vec) * 100 else NA_real_

  data.frame(
    Fold              = x$fold_id,
    Outer_Converged   = isTRUE(x$gmert_fit$converged_out),
    Inner_Conv_Percent= round(inner_pct, 1),
    F1_GMERT          = round(x$metrics$f1_gmert, 4),
    F1_GLMM           = round(x$metrics$f1_glmm, 4),
    check.names = FALSE
  )
}))

# sort by fold, just in case
fold_table <- fold_table[order(fold_table$Fold), ]

# pretty print
knitr::kable(
  fold_table,
  align = c("c","c","r","r","r"),
  caption = "Per-fold convergence status and F1 scores (GMERT vs GLMM)"
)

cv_summary

```

The 10-fold cross-validation indicates that the GMERT algorithm achieves stable convergence in its inner EM loop (92–100 % across folds) but fails to meet the strict outer PQL tolerance in all folds. This pattern suggests that the pseudo-likelihood updates stabilize numerically without fully satisfying the stopping rule.

In terms of predictive performance, GMERT attains an average accuracy of 0.76 and an F1-score of 0.82, while the benchmark GLMM reaches 0.79 and 0.84, respectively. The GLMM therefore performs slightly better on this dataset, which is expected given that the data were generated from a model with an approximately linear fixed component and with few predictors. The small performance gap and the higher variability observed for GMERT reflect the flexibility and stochastic nature of its tree-based fixed-effects estimation.

Overall, the results confirm that the GMERT procedure is numerically stable and competitive with the parametric GLMM, though further tuning of tree complexity and convergence thresholds could improve outer-loop stability and narrow the performance gap.

```{r, message=FALSE, warning=FALSE}
# build a compact data frame with F1 per fold
f1_df <- do.call(rbind, lapply(cv_results, function(x) {
  data.frame(
    Fold = factor(x$fold_id),
    Model = rep(c("GMERT","GLMM"), each = 1),
    F1 = c(x$metrics$f1_gmert, x$metrics$f1_glmm)
  )
}))

# long format ensures one bar per model per fold
ggplot(f1_df, aes(x = Fold, y = F1, fill = Model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", F1)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = c("#4E79A7", "#E15759")) +
  labs(
    title = "Fold-wise F1 Scores: GMERT vs GLMM",
    subtitle = "10-fold cluster-balanced cross-validation",
    x = "Fold",
    y = "F1 Score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11)
  ) +
  ylim(0, 1)
```
















