---
title: "MERT implementation"
author: "Paolo Colussi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
rm(list = ls())
library(rpart)   # CART trees used for the fixed (nonlinear) component f(X)
library(MASS)    # mvrnorm() to simulate correlated random effects for clusters
```

## MERT

The MERT model is defined as:

\[
y_i = f(X_i) + Z_i b_i + \epsilon_i, \\
b_i \sim N(0,D), \epsilon_i \sim N(0,R_i), \\
i = 1, . . . , n,
\]

where $y_i = [y_{i1} , . . . , y_{in_i} ]^T$ is the $n_i × 1$ vector of responses for the $n_i$ observations in cluster $i$, $X_i = [x_{i1} , . . . , x_{in_i} ]^T$ is the $n_i × p$ matrix of fixed-effects covariates, $Z_i = [z_{i1}, . . . , z_{in_i} ]^T$ is the $n_i × q$ matrix of random-effects covariates, $\epsilon_i = [\epsilon_{i1} , . . . , \epsilon_{in_i}]^T$ is the $n_i × 1$ vector of errors, $b_i = (b_{i1} , . . . , b_{iq} )^T$ is the $q × 1$ unknown vector of random effects for cluster $i$. $f(X_i)$ is the fixed effect part for the multilevel model, in the LME model is equal to $X_i\beta$, where $\beta$ is p × 1 unknown vector of parameters for the fixed effects. In our case we will focus on regression trees, but you can actually use any kind of ML or non ML techniques.  
The total number of observations is $N = \sum^n_{i=1}n_i$ . The covariance matrix of $b_i$ is $D$ while $R_i$ is the covariance matrix of $\epsilon_i$. We also assume also assumes that $b_i$ and $\epsilon_i$ are independent and normally distributed and that the between-clusters observations are independent (this is an usual assumption in LME models). Hence, the covariance matrix of the vector of observations $y_i$ in cluster $i$ is $V_i = Cov(y_i) = Z_i D Z_i^T + R_i$ , and $V = Cov(y) = diag(V_1 , . . . , V_n)$, where $y = [y^T_1 , . . . , y^T_n ]^T$ . We will further assume that the correlation is induced solely via the between-clusters variation, that is, $R_i$ is diagonal ($R_i = \sigma^2I_{n_i} , i = 1, . . . , n$). 

### Algorithm

The MERT algorithm is the ML-based EM-algorithm in which we replace the linear structure used to estimate the fixed part of the model by a standard tree structure. The algorithm is as follows:

**Step 0**

Set $r = 0$. Let $\hat b_{i(0)} = 0$, $\hat\sigma^2_{(0)} = 1$, and $\hat D_{(0)} = I_q$.

**Step 1**

Set $r = r + 1$. 

Update $y^*_{i(r)}$, $\hat f(X_i)_{(r)}$ , and $\hat b_{i(r)}$

$(i)$
\[
y^*_{i(r)} = y_i - Z_i\hat b_{i(r-1)}, i = 1, . . . , n \\
\]

$(ii)$

Let  $\hat f(X_i)_{(r)}$ an estimate of $f(X_i)$ obtained from a standard tree algorithm with $y^*_{i(r)}$ as responses and $X_i, i = 1, . . . , n$, as covariates. Note that the tree is built as usual using all N individual observations as inputs along with their covariate vectors

$(iii)$
\[
\hat b_{i(r)} = \hat D_{(r-1)} Z_i^T \hat V_{i(r-1)}^{-1} (y_i - \hat f(X_i)_{(r)}), i = 1,. . .,n
\]

In R:
```{r b_fun}
# b_fun: updates random effects b_i given current D, V_i, and residuals (y - fhat)
b_fun <- function(G,    # G   : number of clusters
                  Z,    # Z   : N x q random-effects design (RIC => q = 2)
                  Vi,   # Vi  : list of length G with V_i matrices (n_i x n_i)
                  D,    # D   : q x q covariance of random effects
                  idx,  # idx : list of length G with row indices for each cluster
                  y,    # y   : response vector length N
                  fhat # target matrix
                  ){
  b <- matrix(0,G,ncol(Z))
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]                      # cluster-specific Z_i
    rhs <- y[idx[[g]]] - fhat[idx[[g]]]                    # residual y - fhat on cluster i
    b[g, ] <- (D %*% t(Zi) %*% solve(Vi[[g]]) %*% rhs)       # b_i = D Z_i^T V_i^{-1} (y - fhat)
  }
  b
}
```

where $\hat V_{i(r-1)} = Z_i \hat D_{(r-1)} Z_i^T + \hat \sigma^2_{(r-1)}I_{n_i}, i=1,...,n$

In R:
```{r Vi_fun}
# Vi_fun: builds list of V_i = Z_i D Z_i^T + sigma^2 I_{n_i}
Vi_fun <- function(Z,   # Z   : N x q random-effects design
                   D,   # D   : q x q covariance of random effects
                   s2,  # s2  : scalar sigma^2 (residual variance; homoskedastic)
                   G,   # G   : number of clusters
                   idx  # idx : list of row indices per cluster
                   ){
  Vi <- list()
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]                      # Z_i
    Vi[[g]] <- Zi %*% D %*% t(Zi) + as.numeric(s2) * diag(nrow(Zi))  # V_i
  }
  # Output:
  # list of length G with n_i x n_i covariance matrices V_i
  return(Vi)
}
```

**Step 2**

Update $\hat \sigma^2_{(r)}$ and $\hat D_{(r)}$ using

$(i)$
\[
\hat \sigma^2_{(r)} = N^{-1} \sum^n_{i=1} {\hat\epsilon^T_{i(r)} \hat \epsilon_{i(r)} + \hat\sigma^2_{(r-1)}[n_i - \hat\sigma^2_{(r-1)}trace(\hat V_{i(r-1)})]}
\]

where $\hat \epsilon_{i(r)} = y_i - \hat f(X_i)_{(r)} - Z_i \hat b_{i(r)}$

In R:
```{r sigma_fun}
# sigma_fun: updates sigma^2 using paper's closed-form expression
sigma_fun <- function(N,   # N   : total number of observations
                      G,   # G   : number of clusters
                      idx, # idx : list of row indices per cluster
                      b,   # b   : matrix q × n vector of random effects 
                      y,   # y   : response vector length N
                      Z,   # Z   : N x q random-effects design
                      D,   # D   : q x q covariance of random effects
                      Vi,  # Vi  : list of length G with V_i matrices (n_i x n_i)
                      s2,  # s2  : scalar sigma^2 (residual variance; homoskedastic)
                      fhat # fhat: current fixed-effect fit
                      ){
  term_sum <- 0
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]                          
    e <- y[idx[[g]]] - fhat[idx[[g]]] - Zi%*% b[g, ]       # residuals epsilon_i = y - fhat - Z_i b_i
    term_sum <- term_sum + crossprod(e) + s2 * (length(idx[[g]]) -
                       s2 * sum(diag(solve(Vi[[g]]))))     # add cluster-i contribution
  }
  # Output:
  #   scalar sigma^2 new estimate
  term_sum / N                                             # divide by total N
}

```

$(i)$
\[
\hat D_{(r)} = n^{-1} \sum^n_{i=1} {\hat b_{i(r)} \hat b^T_{i(r)} + [\hat\ D_{(r-1)} - \hat\ D_{(r-1)} Z_i^T \hat V^{-1}_{i(r-1)} Z_i \hat\ D_{(r-1)}]}
\]

In R:
```{r D_fun}
# D_fun: updates D using the paper's expression (average of b_i b_i^T plus shrinkage term)
D_fun <- function(G,   # G   : number of clusters
                  idx, # idx : list of row indices per cluster
                  b,   # b   : matrix q × n vector of random effects 
                  Z,   # Z   : N x q random-effects design
                  D,   # D   : q x q covariance of random effects
                  Vi  # Vi  : list of length G with V_i matrices (n_i x n_i)
                  ){
  q <- ncol(Z)
  S <- matrix(0, q, q)                                     # accumulator
  for (g in seq_len(G)) {
    Zi <- Z[idx[[g]], , drop = FALSE]                      # Z_i
    S <- S + tcrossprod(b[g, ]) +                          # b_i b_i^T
      (D - D %*% t(Zi) %*% solve(Vi[[g]]) %*% Zi %*% D)    # shrinkage term per cluster
  }
  # Output:
  #   q x q updated D
  S/G                                                      # average over clusters
}
```

**Step 3**

Repeat steps 1 and 2 until convergence

The convergence is monitored by computing, at each iteration, the following generalized log-likelihood (GLL) criterion:

\[
GLL(f, b_i | y) = \sum^n_{i=1} {[y_i - f(X_i) - Z_i b_i]^T R_i^{-1} [y_i - f(X_i) - Z_i b_i] + b_i^T D^{-1} b_i + log|D| + log|R_i|}
\]

In R:

```{r gll_fun}

# gll_fun: computes generalized log-likelihood (up to additive constant)
gll_fun <- function(idx, # idx : list of row indices per cluster
                    b,   # b   : matrix q × n vector of random effects 
                    y,   # y   : response vector length N
                    Z,   # Z   : N x q random-effects design
                    D,   # D   : q x q covariance of random effects
                    s2,  # s2  : scalar sigma^2 (residual variance; homoskedastic)
                    fhat # fhat: current fixed-effect fit
                    ){
  logdetD <- as.numeric(determinant(D, logarithm = TRUE)$modulus)     # log|D|
  term_b <- sum(rowSums((b %*% solve(D)) * b))                         # sum_i b_i^T D^{-1} b_i
  term_r <- 0
  term_logRi <- 0
  for (i in seq_along(idx)) {
    ri <- y[idx[[i]]] - fhat[idx[[i]]]                                 # start residual y - fhat
           - as.numeric(Z[idx[[i]], , drop = FALSE] %*% b[i, ])        # minus Z_i b_i
    term_r <- term_r + sum(ri^2) / s2                                  # (1/s2) * ||ri||^2   (note: ˆ is ^)
    term_logRi <- term_logRi + length(idx[[i]]) * log(s2)              # n_i * log(s2)
  }
  
  # Output:
  #   scalar GLL value (up to constant)
  # GLL = sum_i [ (y - f - Zb)^T R_i^{-1} (y - f - Zb) + log|R_i| ] +
  # + sum_i b_i^T D^{-1} b_i + G*log|D|
  term_r + term_logRi + term_b + length(idx) * logdetD                 # assemble GLL
}
```

To check convergence we use the Aitken stop, *this is not in the paper*, and I'm really unsure if it good enough, but looks like it 

In R:

```{r ait_stop}
# ait_stop: Aitken acceleration based stopping rule
ait_stop <- function(gll,  #gll : numeric vector with GLL history (needs last three values)
                     tol,  #tol : tolerance for relative error to asymptote
                     it    #it  : current iteration index 
                     ){
  converged <- FALSE
  l1 <- gll[it + 2] - gll[it + 1]                                      # L_t
  l2 <- gll[it + 1] - gll[it]                                          # L_{t-1}
  a <- l1 / l2                                                         # Aitken ratio
  if (is.finite(a) && a > 0 && a < 1) {                                # only meaningful if 0<a<1
    L_inf <- gll[it + 1] + l1 / (1 - a)                                # extrapolate asymptote
    err <- abs(L_inf - gll[it + 2]) / (abs(L_inf) + 1e-12)             # relative error to asymptote
    if (err < tol) converged <- TRUE                                   # declare convergence
  }
  converged                                                            # return flag
}
```

### Implementation

Here we put the algorithm in a single function with many parameter to be able to control the convergence and to be able to tune the tree from step 1. Also we have a convergence plot for the GLL, if we want to check visually the convergence. 

```{r fit_fun}
fit_mert_RIC <- function(df,               # df: data.frame with columns
                                           #  - id: cluster identifier (factor or integer)
                                           #  - y: numeric response
                                           #  - x1,x2,x3: numeric covariates
                                           #  - f/leaf: true generating leaf
                         max_iter = 1000,  # max_iter: maximum number of EM iterations
                         tol = 1e-6,       # tol: convergence tolerance (for Aitken acceleration on GLL)
                         cp = 0.0,         # cp: rpart complexity parameter (pruning threshold)
                         minsplit = 20,    # minsplit: minimum number of obs that must exist in a node to split
                         xval = 10,        # xval: number of cross-validation folds used inside rpart
                         plot_convergence = FALSE # plot_convergence: if TRUE, plots trajectory of GLL vs iteration
                         ) {

  # --- Basic setup ---
  N <- nrow(df)                                        # total number of rows
  G <- length(sort(unique(df$id)))                     # number of clusters
  idx_by_cluster <- split(seq_len(N), df$id)           # list: indices for each cluster

  y <- df$y                                           # response vector
  Z <- cbind(1, df$x1)                                # random effects design: intercept + slope on x1
  q <- ncol(Z)                                        # dimension of random effects (q=2)

  # --- Initialization (Step 0 of EM) ---
  sigma2 <- 1                                         # initialize residual variance
  D <- diag(2)                                        # initialize random-effects covariance as identity
  b <- matrix(0, G, q)                                # initialize all cluster random effects at zero
  gll <- c(0, 0)                                      # store GLL values (two placeholders for Aitken)
  it <- 0L                                            # iteration counter

  # --- EM algorithm loop ---
  repeat {
    it <- it + 1L

    # (1.i) Partial E-step: compute y* = y - Z b
    zb <- numeric(N)                                  # vector to hold cluster random effect contributions
    for (g in seq_len(G)) {
      idx <- idx_by_cluster[[g]]                      # rows belonging to cluster g
      zb[idx] <- Z[idx, , drop = FALSE] %*% b[g, ]    # compute Z_i b_i for this cluster
    }
    y_star <- y - zb                                  # pseudo-response for fixed effects tree

    # (1.ii) Fit regression tree to y* to update f(X)
    ctrl <- rpart.control(cp = cp, minsplit = minsplit, xval = xval)  # tree growth control
    tree <- rpart(y_star ~ x1 + x2 + x3, data = df,
                  method = "anova", control = ctrl)   # regression tree on pseudo-response

    fhat <- as.numeric(predict(tree, df[, -2]))       # fitted fixed part values for all obs

    # (1.iii) Update random effects b_i
    Vi <- Vi_fun(Z = Z, D = D, s2 = sigma2, G = G, idx = idx_by_cluster)  # compute cluster covariance matrices
    b  <- b_fun(G = G, Z = Z, Vi = Vi, D = D,idx = idx_by_cluster, 
                y = y, fhat = fhat)          # update b_i for each cluster

    # (2.i) Update sigma^2 (residual variance)
    sigma2 <- sigma_fun(N = N, G = G, idx = idx_by_cluster,
                        b = b, y = y, Z = Z, D = D,
                        Vi = Vi, s2 = sigma2, fhat = fhat)

    # (2.ii) Update D (covariance of random effects)
    D <- D_fun(G = G, idx = idx_by_cluster,
               Z = Z, D = D, Vi = Vi, b = b)

    # --- Convergence monitoring ---
    gll[it+2] <- gll_fun(D = D, b = b, idx = idx_by_cluster,
                         Z = Z, y = y, fhat = fhat, s2 = sigma2)
    # Append GLL value for this iteration. The index shift (+2) leaves room for Aitken method.

    if (ait_stop(gll = gll, it = it, tol = tol)) {    # check Aitken acceleration criterion
      n_iter <- it
      converged <- TRUE
      break                                           # stop EM loop if converged
    }

    if (it >= max_iter) {                             # guard against infinite loop
      n_iter <- it
      converged <- FALSE
      message(sprintf("WARNING: did not converge in %d iterations.", max_iter))
      break
    }
  }

  # --- Optional convergence plot ---
  if (isTRUE(plot_convergence)) {
    op <- par(no.readonly = TRUE); on.exit(par(op), add = TRUE)
    if (length(gll) >= 3) {                           # enough GLL values to plot
      g_idx <- 3:length(gll)                          # skip placeholders
      it_axis <- seq_along(g_idx)                     # iterations on x-axis
      plot(it_axis, gll[g_idx], type = "l", lwd = 2,
           xlab = "Iteration", ylab = "GLL (up to const)",
           main = "EM Convergence (GLL)")
      if (!is.na(n_iter)) {                           # mark stopping iteration
        abline(v = n_iter, lty = 2, col = "red")
        mtext(sprintf("stop @ %d", n_iter),
              side = 3, line = 0.2, col = "red")
      }
    }
  }

  # --- Return results ---
  list(
    tree      = tree,       # fitted rpart tree (fixed-effects function f(X))
    b         = b,          # estimated random effects for each cluster (matrix G x q)
    D         = D,          # estimated covariance matrix of random effects
    sigma2    = sigma2,     # estimated residual variance
    converged = converged,  # logical flag whether converged
    n_iter    = n_iter,     # number of iterations actually run
    train_ids = df$id,      # training cluster identifiers
    gll       = gll,        # sequence of GLL values (for diagnostics)
    tol       = tol         # tolerance value used for stopping
  )
}


```

### Prediction

To predict the response for a new observation that belongs to a cluster among those used to fit the MERT model, we use both its corresponding population-averaged tree prediction and the predicted random part corresponding to its cluster. For a new observation that belongs to a cluster not included in the sample used to estimate the model parameters, we can only take the corresponding population-averaged tree prediction.

```{r}
predict_mert_RIC <- function(fit, new_df) {
  # fixed part: regression tree predictions using predictors only
  fhat <- as.numeric(predict(fit$tree, new_df))        # tree fitted on y*; use on new data
  # construct random effects design for new data: [1, x1]
  Znew <- cbind(1, new_df$x1)
  # container for random effects contributions
  add <- numeric(nrow(new_df))
  # clusters seen during training (so we have estimated b_i)
  clus_fit <- sort(unique(fit$train_ids))
  # map each new cluster id to its index in clus_fit (NA if unseen)
  map <- match(new_df$id, clus_fit)
  seen <- !is.na(map)                                  # TRUE for rows belonging to seen clusters
  if (any(seen)) {
    # add random effect contribution: row-wise sum of [1, x1] * b_i
    add[seen] <- rowSums(Znew[seen, , drop = FALSE] *
                         fit$b[map[seen], ])
  }
  # total prediction = fixed part + random effects contribution
  fhat + add
}
```


## Cross-validation implementation

To make our model more robust, we perform cross-validation of the MERT model at the cluster level. Each fold trains on $K-1$ sets of clusters and tests on the held-out set, so the model is evaluated on completely unseen clusters (where random effects can’t be estimated, so *I should probably change that*).

For each fold the code computes:

* PMSE (prediction mean squared error) on the test set.
* Number of leaves in the fitted tree (a measure of tree complexity).
* Convergence flag for the EM algorithm.
* The full fit object, so you can inspect estimated parameters.
* Leaf assignment accuracy:
  - On training data: percentage of observations assigned to their true generating leaf when each terminal node is labeled by its majority true leaf.
  - On test data: same percentage, for held-out clusters.
* An overall total percentage of “right trees” across all folds, computed by pooling all test rows.

Two versions are provided:

* `cv10_mert_RIC`: sequential, runs folds one after another.
* `cv10_mert_RIC_parallel`: parallelized with parLapply for faster evaluation on multi-core machines.

We also need a function to assign a plausible terminal node for the new data that we predict with the tree. To do so, we use the leaf mean (yval) for the tree that we trained. Then, we map each predicted value back to the leaf whose yval matches, and return the ID of the leaf of the trained tree.

```{r leaf_where}
# Helper: recover terminal node IDs for NEW data in an rpart regression tree
leaf_per <- function(tree, train, newdata, digits = 12) {
        # Map each learned terminal node (from training) to its majority TRUE leaf
  node_train <- tree$where                             # node ID per TRAIN row
  tab <- table(node = node_train, leaf = train$leaf)       # cross-tab nodes vs true leaves
  major_by_node <- apply(tab, 1, which.max)                # majority true leaf per node (1..4)
  pred_leaf_train <- unname(major_by_node[as.character(node_train)])
  fr <- tree$frame                                           # rpart frame (nodes metadata)
  leaf_rows <- fr$var == "<leaf>"                            # logical: which rows are leaves
  leaf_ids  <- as.integer(row.names(fr)[leaf_rows])          # numeric node IDs for leaves
  yvals     <- fr$yval[leaf_rows]                            # leaf means
  yhat <- as.numeric(predict(tree, newdata))                 # predicted leaf means for newdata
  node_test <- match(round(yhat, digits), round(yvals, digits))    # map each yhat to a leaf yval
  tab <- table(node = leaf_ids[node_test], leaf = newdata$leaf)       # cross-tab nodes vs true leaves
  major_by_node <- apply(tab, 1, which.max)
  pred_leaf_test <- unname(major_by_node[as.character(leaf_ids[node_test])])
  leaf_acc <- mean(c(pred_leaf_train,pred_leaf_test) == c(train$leaf,newdata$leaf)) * 100 
  leaf_acc
 }
```

Now we can define our functions, first he sequential one

```{r cross-validation}

# Sequential K-fold CV at the cluster level
cv10_mert_RIC <- function(df, K = 10, seed = 123,
                          max_iter = 1000, tol = 1e-6,
                          cp = 0.0, minsplit = 20, xval = 10) {
  set.seed(seed)                                             # reproducibility for fold assignment
  clus <- sort(unique(df$id))                                # unique cluster IDs
  K <- min(K, length(clus))                                  # if clusters < K, cap K
  folds <- split(clus, sample(rep(seq_len(K),                # assign clusters to folds uniformly
                                  length.out = length(clus))))

  pmse   <- numeric(K)                                       # PMSE per fold (test)
  leaves <- integer(K)                                       # number of terminal nodes per fitted tree
  conv   <- logical(K)                                       # EM convergence flag per fold

  fits <- vector("list", K)                                  # store full fit objects (inspect later)
  leaf_acc_train <- numeric(K)                               # % correct leaf assignment on training data
  leaf_acc_test  <- numeric(K)                               # % correct leaf assignment on test data
  n_test_vec     <- integer(K)                               # number of test rows per fold
  correct_test_counts <- integer(K)                          # count of correct leaf assignments on test

  for (k in seq_len(K)) {
    test_ids <- folds[[k]]                                   # held-out clusters for this fold
    train <- df[!df$id %in% test_ids, , drop = FALSE]        # training subset (all other clusters)
    test  <- df[ df$id %in% test_ids, , drop = FALSE]        # test subset (held-out clusters only)

    fit <- fit_mert_RIC(train, max_iter = max_iter, tol = tol,
                        cp = cp, minsplit = minsplit, xval = xval)   # fit model on training clusters
    fits[[k]] <- fit                                         # store the whole fit

    yhat <- predict_mert_RIC(fit, test)                      # predict on test rows (new clusters => fixed part only)
    pmse[k]   <- mean((test$y - yhat)^2)                     # test PMSE
    leaves[k] <- length(unique(fit$tree$where))              # tree complexity (number of unique terminal nodes)
    conv[k]   <- isTRUE(fit$converged)                       # did EM converge?



    # Aggregate counts to compute a global test percentage across folds
    n_test_vec[k] <- nrow(test)
    correct_test_counts[k] <- sum(pred_leaf_test[ok] == test$leaf[ok])

    # Optional console log per fold (comment out if too chatty)
    cat(sprintf("[Fold %d/%d] PMSE=%.4f | leaves=%d | converged=%s | leaf_acc_train=%.2f%% | leaf_acc_test=%.2f%%\n",
                k, K, pmse[k], leaves[k], conv[k], leaf_acc_train[k], leaf_acc_test[k]))
  }

  total_right_trees_pct <- 100 * sum(correct_test_counts) / sum(n_test_vec)  # observation-weighted over all test rows

  list(
    PMSE_mean              = mean(pmse),                  # average test PMSE across folds
    PMSE_sd                = sd(pmse),                    # std dev of test PMSE
    PMSE_by_fold           = pmse,                        # vector of test PMSE per fold
    avg_leaves             = mean(leaves),                # average number of leaves
    converged_by_fold      = conv,                        # logical vector of EM convergence flags
    fits                   = fits,                        # list of full fit objects (one per fold)
    leaf_acc_train_by_fold = leaf_acc_train,              # % correct leaves on training per fold
    leaf_acc_test_by_fold  = leaf_acc_test,               # % correct leaves on test per fold
    total_right_trees_pct  = total_right_trees_pct        # pooled percentage of correct test leaves
  )
}
```

And then we define the parallel version

```{r cross-validation_par}
# Parallel K-fold CV at the cluster level (PSOCK cluster; portable)
cv10_mert_RIC_parallel <- function(df, K = 10, seed = 123,
                                   max_iter = 1000, tol = 1e-6,
                                   cp = 0.0, minsplit = 20, xval = 10,
                                   n_cores = NULL) {
  set.seed(seed)                                             # reproducibility
  clus <- sort(unique(df$id))                                # unique cluster IDs
  K <- min(K, length(clus))                                  # cap K if needed
  folds <- split(clus, sample(rep(seq_len(K),                # assign clusters to folds
                                  length.out = length(clus))))

  if (is.null(n_cores)) {                                    # choose core count: <= K and <= physical cores
    n_cores <- max(1, min(K, parallel::detectCores(logical = FALSE)))
  }
  cl <- parallel::makeCluster(n_cores)                       # start cluster 
  on.exit(parallel::stopCluster(cl), add = TRUE)             # ensure cleanup

  parallel::clusterSetRNGStream(cl, seed)                    # synchronized RNG streams per worker
  parallel::clusterEvalQ(cl, { library(rpart); library(MASS) })  # load packages
  parallel::clusterExport(                                   # export needed objects/functions 
    cl,
    varlist = c("df","folds","K","max_iter","tol","cp","minsplit","xval",
                "fit_mert_RIC","predict_mert_RIC","leaf_where","Vi_fun",
                "b_fun","D_fun","sigma_fun","gll_fun","ait_stop"),
    envir = environment()
  )

  # Run each fold on a worker
  res_list <- parallel::parLapply(cl, seq_len(K), function(k) {
    tryCatch({
    test_ids <- folds[[k]]                                   # held-out clusters
    train <- df[!df$id %in% test_ids, , drop = FALSE]        # training subset
    test  <- df[ df$id %in% test_ids, , drop = FALSE]        # test subset

    fit <- fit_mert_RIC(train, max_iter = max_iter, tol = tol,
                        cp = cp, minsplit = minsplit, xval = xval)   # train on worker

    yhat <- predict_mert_RIC(fit, test)                      # predict on test rows

    pmse <- mean((test$y - yhat)^2)                          # test PMSE
    leaves <- length(unique(fit$tree$where))                 # tree size
    converged <- isTRUE(fit$converged)                       # EM convergence


    leaf_acc <- leaf_per(fit$tree,train,test)

    list(
      fit        = fit,                                  # full fitted model for this fold
      pmse       = pmse,                                 # test PMSE
      leaves     = leaves,                               # number of leaves
      converged  = converged,                            # EM convergence flag
      leaf_acc   = leaf_acc,
      n_test     = nrow(test),                           # number of test rows
      correct_test_count = correct_test_count            # count of correct test leaves
    )
    })
  })

  bad <- vapply(res_list, function(x) identical(x$.ok, FALSE), logical(1))
if (any(bad)) {
  flds <- vapply(res_list[bad], `[[`, integer(1), "fold")
  msgs <- vapply(res_list[bad], `[[`, character(1), "message")
  stop(sprintf("Parallel CV failed in folds %s; first error: %s",
               paste(flds, collapse = ", "), msgs[1]))
}
  
  # Aggregate results back on the master
  pmse    <- vapply(res_list, `[[`, numeric(1), "pmse")                # numeric vector length K
  leaves  <- vapply(res_list, `[[`, integer(1), "leaves")
  conv    <- vapply(res_list, `[[`, logical(1), "converged")
  fits    <- lapply(res_list, `[[`, "fit")
  leaf_acc_train <- vapply(res_list, `[[`, numeric(1), "leaf_acc_train")
  leaf_acc_test  <- vapply(res_list, `[[`, numeric(1), "leaf_acc_test")
  n_test_vec     <- vapply(res_list, `[[`, integer(1), "n_test")
  correct_counts <- vapply(res_list, `[[`, integer(1), "correct_test_count")

  # console log per fold (originally meant only for the sequential function, but useful to see immediatly how many folds converged)
  for (k in seq_len(K)) {
    cat(sprintf("[Fold %d/%d] PMSE=%.4f | leaves=%d | converged=%s | leaf_acc_train=%.2f%% | leaf_acc_test=%.2f%%\n",
                k, K, pmse[k], leaves[k], conv[k], leaf_acc_train[k], leaf_acc_test[k]))
  }

  total_right_trees_pct <- 100 * sum(correct_counts) / sum(n_test_vec) # pooled percentage over all test rows

  list(
    PMSE_mean             = mean(pmse),                 # average test PMSE
    PMSE_sd               = sd(pmse),                   # dispersion of test PMSE
    PMSE_by_fold          = pmse,                       # per-fold PMSE
    avg_leaves            = mean(leaves),               # average tree size
    converged_by_fold     = conv,                       # per-fold EM convergence flags
    fits                  = fits,                       # per-fold fitted models
    leaf_acc_by_fold= leaf_acc,
    total_right_trees_pct = total_right_trees_pct,      # pooled correct-leaf percentage on test
    n_cores_used          = n_cores                     # number of worker processes used
  )
}

```

## Simulation

This function simulate the data in the same way the paper does, using a tree for determine the data generating process of $y$.

```{r}
# --- Simulate clustered data with 4-leaf fixed tree and RIC random effects ---
# Purpose: create a synthetic dataset with cluster IDs, covariates, outcome y, and the
#          TRUE leaf label used by the data-generating partition. This lets us later
#          evaluate whether the learned tree recovers the correct partition.
simulate_data_paper <- function(n_clusters = 100, n_per_cluster = 55,
                              mu = c(-20, -10, 10, 20),
                              D = matrix(c(0.25, 0.00, 0.00, 0.25), 2, 2),
                              sigma2 = 1) {
  N  <- n_clusters * n_per_cluster
  id <- rep(seq_len(n_clusters), each = n_per_cluster)
  x1 <- runif(N, 0, 10)
  x2 <- runif(N, 0, 10)
  x3 <- runif(N, 0, 10)
  
  # True 4-region partition (axis-aligned). We'll check if the learned tree matches this.
  leaf <- ifelse(x1 <= 5 & x2 <= 5, 1,
                 ifelse(x1 <= 5 & x2 > 5, 2,
                        ifelse(x1 >  5 & x3 <= 5, 3, 4)))
  f_fixed <- mu[leaf]
  
  # Cluster random effects b_i ~ N(0, D), applied to [1, x1]
  B <- MASS::mvrnorm(n = n_clusters, mu = c(0, 0), Sigma = D)  # G x 2
  Z <- cbind(1, x1)
  rand <- rowSums(Z * B[id, ])
  
  # Outcome with residual noise
  y <- f_fixed + rand + rnorm(N, 0, sqrt(sigma2))
  data.frame(id, y, x1, x2, x3, leaf)
}
```

## Simulation like in the 

```{r}
simulation_paper <- function(
  n_rep = 100,
  n_clusters = 100,
  n_per_cluster = 55,
  mu = c(-20, -10, 10, 20),
  D = matrix(c(0.25, 0.00, 0.00, 0.25), 2, 2),
  sigma2 = 1,
  seed = 123,
  max_iter = 1000,
  tol = 1e-6,
  cp = 0.0,
  minsplit = 20,
  xval = 10,
  n_workers = 10,
  use_parallel = TRUE
) {
  # --- basic checks -----------------------------------------------------------
  if (n_per_cluster < 5L) {
    stop("n_per_cluster must be >= 5 to sample 5 train obs per cluster.")
  }

  # Master seed ensures reproducibility of: replicate seeds + any serial path.
  set.seed(seed)

  # Create a reproducible seed per replicate (so parallel order doesn't matter).
  rep_seeds <- sample.int(1000000, n_rep)

  # Worker function: runs ONE replicate safely and returns a named list
  one_rep <- function(rep_id, rep_seed) {
    set.seed(rep_seed)

    # 1) Simulate data
    data_ric <- simulate_data_paper(n_clusters, n_per_cluster, mu, D, sigma2)

    # 2) Stratified split: 5 train obs from each cluster id, rest to test
    ids <- data_ric$id
    train_idx <- unlist(
      tapply(seq_len(nrow(data_ric)), ids, function(idx) sample(idx, 5L))
    )
    train <- data_ric[train_idx, , drop = FALSE]
    test  <- data_ric[-train_idx, , drop = FALSE]

    # 3) Fit model with protection against failures (keep the run going)
    fit <- fit_mert_RIC(train, max_iter = max_iter, tol = tol,
                   cp = cp, minsplit = minsplit, xval = xval)

    # 4) Predict + metrics
    yhat <- predict_mert_RIC(fit, test)
    pmse <- mean((test$y - yhat)^2)

    # custom leaf metric on train/test, if available
    lper <- leaf_per(fit$tree, train, test)

    list(
      rep = rep_id,
      pmse = pmse,
      converged = isTRUE(fit$converged),
      leaves_per = as.numeric(lper),
      seed = rep_seed
    )
  }

  # --- run replicates (parallel or serial) -----------------------------------
  results_list <- NULL

  if (use_parallel && n_workers > 1L) {
    cl <- parallel::makeCluster(n_workers)
    on.exit(try(parallel::stopCluster(cl), silent = TRUE), add = TRUE)

    # Export required objects/functions into workers’ environments
    parallel::clusterEvalQ(cl, { library(rpart); library(MASS) })
    parallel::clusterExport(
      cl,
      varlist = c("simulate_data_paper", "fit_mert_RIC", "predict_mert_RIC",
                  "leaf_per", "n_clusters", "n_per_cluster", "mu", "D",
                  "sigma2", "max_iter", "tol", "cp", "minsplit", "xval","Vi_fun",
                "b_fun","D_fun","sigma_fun","gll_fun","ait_stop"),
      envir = environment()
    )

    # Ensure packages (if any) used by your functions are attached on workers
    # parallel::clusterEvalQ(cl, { library(yourPkgIfNeeded) })

    # Build args and dispatch
    idx <- seq_len(n_rep)
    results_list <- parallel::parLapply(
      cl, idx,
      function(i) {
        one_rep(i, rep_seed = rep_seeds[i])
      }
    )
  } else {
    # Serial fallback
    results_list <- lapply(seq_len(n_rep), function(i) one_rep(i, rep_seeds[i]))
  }

  # --- aggregate --------------------------------------------------------------
  # Bind to data.frame
  res <- do.call(rbind, lapply(results_list, as.data.frame))

  # Summaries with NA-robust stats
  pmse_vec <- res$pmse
  leaves_per_vec <- res$leaves_per

  out <- list(
    avg_tree = mean(leaves_per_vec, na.rm = TRUE),
    avg_pmse = mean(pmse_vec, na.rm = TRUE),
    min_pmse = min(pmse_vec, na.rm = TRUE),
    max_pmse = max(pmse_vec, na.rm = TRUE),
    sd_pmse  = stats::sd(pmse_vec, na.rm = TRUE),
    med_pmse = stats::median(pmse_vec, na.rm = TRUE),
    conv     = res$converged,              # vector per-replicate (TRUE/FALSE)
    converged_ratio = mean(res$converged, na.rm = TRUE),
    per_rep  = res                          # full per-replicate table
  )

  return(out)
}

```


## Playground

```{r playgroung, eval=FALSE}

# dat <- simulate_data_paper(n_clusters = 100, n_per_cluster = 55,
#                          mu = c(-20, -10, 10, 20),
#                          D = matrix(c(0.25, 0, 0, 0.25), 2, 2),
#                          sigma2 = 1, seed = 42)

# cv_res <- cv10_mert_RIC_parallel(dat, seed = 7, max_iter = 1000, tol = 1e-10,
#                                  cp = 0.0, minsplit = 50, xval = 10, n_cores = 10)

# cv_res <- cv10_mert_RIC(dat, seed = 7, max_iter = 100, tol = 1e-35,
#                                  cp = 0.0, minsplit = 50, xval = 10)
# print(cv_res)
# cat(sprintf("Total right trees on CV test sets: %.2f%%\n", cv_res$total_right_trees_pct))

#fit <- fit_mert_RIC(dat, max_iter = 1000)

# simulation_paper <- function(n_rep=100, n_clusters = 100, n_per_cluster = 55,
#                             mu = c(-20, -10, 10, 20),
#                             D = matrix(c(0.25, 0.00, 0.00, 0.25), 2, 2),
#                             sigma2 = 1, seed = 123,
#                             max_iter = 1000, tol = 1e-6,
#                             cp = 0.0, minsplit = 20, xval = 10){
### RIC as the paper
#set.seed(seed)
#pmse<- c()
#leaves_per <- c()
#conv <- c()
#for (i in 1:n_rep) {
#data_ric <- simulate_data_paper(n_clusters, n_per_cluster, mu, D, sigma2)
#train_idx <- unlist(tapply(1:5500, data_ric$id, function(idx) sample(idx, 5)))
#train <- data_ric[train_idx,]
#test <- data_ric[-train_idx,]
#fit <- fit_mert_RIC(train, max_iter,tol,cp,minsplit,xval)
#yhat <- predict_mert_RIC(fit, test)   
#pmse[i]   <- mean((test$y - yhat)^2)                     # test PMSE
#conv[i]   <- isTRUE(fit$converged)                       # did EM converged
#leaves_per[i] <- leaf_per(fit$tree,train,test)

#}
#  list(
#    avg_tree <- mean(leaves_per),
#    avg_pmse <- mean(pmse),
#    min_pmse <- min(pmse),
#    max_pmse <- max(pmse),
#    sd_pmse  <- sd(pmse),
#    med_pmse <- median(pmse),
#    conv <- conv
#  )
#}
#


RE_large <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0, 0.00, 0.00, 0), 2, 2))
RE_small <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0, 0.00, 0.00, 0), 2, 2))
RI_large_25sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0.25, 0.00, 0.00, 0), 2, 2))
RI_small_25sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0.25, 0.00, 0.00, 0), 2, 2))
RI_large_50sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0.5, 0.00, 0.00, 0), 2, 2))
RI_small_50sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0.5, 0.00, 0.00, 0), 2, 2)) 
RIC_small_nocorr_25sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0.25, 0.00, 0.00, 0.25), 2, 2))
RIC_large_nocorr_25sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0.25, 0.00, 0.00, 0.25), 2, 2))
RIC_small_nocorr_50sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0.50, 0.00, 0.00, 0.50), 2, 2))
RIC_large_nocorr_50sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0.50, 0.00, 0.00, 0.50), 2, 2))
RIC_small_corr_25sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0.25, 0.125, 0.125, 0.25), 2, 2))
RIC_large_corr_25sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0.25, 0.125, 0.125, 0.25), 2, 2))
RIC_small_corr_50sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(10, 11, 12, 13),D = matrix(c(0.5, 0.25, 0.25, 0.5), 2, 2))
RIC_large_corr_50sd <- simulation_paper(n_rep = 100, n_workers = 10, max_iter = 10000, 
  mu = c(-20, -10, 10, 20),D = matrix(c(0.5, 0.25, 0.25, 0.5), 2, 2))


# List the object names you created
obj_names <- c(
  "RE_large", "RE_small",
  "RI_large_25sd", "RI_large_50sd",
  "RI_small_25sd", "RI_small_50sd",
  "RIC_small_nocorr_25sd", "RIC_small_nocorr_50sd",
  "RIC_large_nocorr_25sd","RIC_large_nocorr_50sd",
  "RIC_small_corr_25sd", "RIC_small_corr_50sd", 
  "RIC_large_corr_25sd",  "RIC_large_corr_50sd"
)

# Helper to extract metrics from one result object by name
.extract_row <- function(name) {
  x <- get(name, envir = .GlobalEnv)
  data.frame(
    model               = name,
    avg_pmse            = x$avg_pmse,
    min_pmse            = x$min_pmse,
    max_pmse            = x$max_pmse,
    sd_pmse             = x$sd_pmse,
    avg_tree_convergence= x$avg_tree,
    convergence_ratio   = x$converged_ratio,
    check.names = FALSE
  )
}

# Build the table
summary_table <- do.call(rbind, lapply(obj_names, .extract_row))

# Optional rounding (set digits as you like)
digits <- 4
num_cols <- c("avg_pmse","min_pmse","max_pmse","sd_pmse",
              "avg_tree_convergence","convergence_ratio")
summary_table[num_cols] <- lapply(summary_table[num_cols], round, digits)

summary_table

```












