---
title: "optimization"
author: "Paolo Colussi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Appendix – Derivation of the Small-System Formulas

One of the things that makes the GMERT algorithm computationally expensive is the need to invert the matrix \(\hat V_i\) of size \(n_i\times n_i\) at each iteration for each group \(i\) for the updates of the random effects \(\hat b_i\), the residual variance \(\hat\sigma^2\), and the random-effects covariance \(\hat D\). What can make this feasible is that the number of random effects \(q\) is typically much smaller than the number of observations per group \(n_i\). Using the Woodbury matrix identity, we can rewrite these updates in terms of inverses of \(q\times q\) matrices instead of \(n_i\times n_i\) matrices, which is computationally much cheaper. This appendix derives these small-system formulas.

### 1. Original GMERT Updates

First, we restate the relevant definitions from the GMERT paper. 
At iteration \(m\), the paper defines the random-effects, residual, and covariance updates as:

\[
\begin{aligned}
\hat b_i^{(m)} &=
\hat D_{(m-1)} (W_i^{1/2(M)} Z_i)^{\!\top}
\hat V_i^{-1(m-1)}
W_i^{1/2(M)}\big(\tilde y_i^{(M)} - \hat f_{(m)}(X_i)\big), \\[4pt]
\hat V_i^{(m-1)} &=
W_i^{1/2(M)} Z_i \hat D_{(m-1)} Z_i^{\!\top} W_i^{1/2(M)}
+ \hat\sigma_{(m-1)}^2 I_{n_i}, \\[6pt]
\hat\sigma^2_{(m)} &=
\frac{1}{N} \sum_{i=1}^{n}
\Big\{
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \hat\sigma^2_{(m-1)}\big[n_i - \hat\sigma^2_{(m-1)} \operatorname{tr}(\hat V_i^{-1(m-1)})\big]
\Big\}, \\[6pt]
\hat D_{(m)} &=
\frac{1}{n}
\sum_{i=1}^n
\Big\{
\hat b_i^{(m)} \hat b_i^{(m)\top}
+
\big[
\hat D_{(m-1)} -
\hat D_{(m-1)}(W_i^{1/2}Z_i)^{\!\top}
\hat V_i^{-1(m-1)} W_i^{1/2}Z_i \hat D_{(m-1)}
\big]
\Big\},
\end{aligned}
\]
with
\[
\hat\epsilon_i^{(m)} = W_i^{1/2}\big(\tilde y_i - \hat f_i - Z_i \hat b_i^{(m)}\big).
\]

All inverses are over the \(n_i\times n_i\) matrices \(\hat V_i\), which dominates computational cost.

---

### 2. Woodbury Identity

To derive small-system expressions, we use the Woodbury matrix identity:
For matrices \(A\in\mathbb{R}^{n_i\times n_i}\), \(U\in\mathbb{R}^{n_i\times q}\),
and \(C\in\mathbb{R}^{q\times q}\),
\[
(A + UCU^{\!\top})^{-1}
= A^{-1} - A^{-1}U\,(C^{-1}+U^{\!\top}A^{-1}U)^{-1}U^{\!\top}A^{-1}.
\]

Set \(A=\sigma^2 I_{n_i}\), \(U=W_i^{1/2}Z_i\), \(C=D\).
Then
\[
\boxed{
V_i^{-1}
= \sigma^{-2}I_{n_i}
- \sigma^{-4}W_i^{1/2}Z_i
\big(D^{-1}+\sigma^{-2}Z_i^{\!\top}W_i Z_i\big)^{-1}
Z_i^{\!\top}W_i^{1/2}.
}
\tag{1}
\]

Define
\[
A_i = D^{-1} + \sigma^{-2} Z_i^{\!\top} W_i Z_i
\quad\text{(\(q\times q\) matrix).}
\]

---

### 3. Small-System Expression for \(b_i\)

Substituting (1) into the paper’s definition of \(\hat b_i\)
and simplifying with \(W_i^{1/2}Z_i=U_i\)
gives

\[
\begin{aligned}
\hat b_i
&= D (W_i^{1/2}Z_i)^{\!\top}
\big[\sigma^{-2}I - \sigma^{-4}W_i^{1/2}Z_iA_i^{-1}Z_i^{\!\top}W_i^{1/2}\big]
W_i^{1/2}(\tilde y_i-\hat f_i) \\[3pt]
&= \big(D^{-1}+\sigma^{-2}Z_i^{\!\top}W_i Z_i\big)^{-1}
      \big(\sigma^{-2}Z_i^{\!\top}W_i(\tilde y_i-\hat f_i)\big).
\end{aligned}
\]

Hence,
\[
\boxed{
\hat b_i =
A_i^{-1}\,\sigma^{-2}Z_i^{\!\top}W_i(\tilde y_i-\hat f_i),
\qquad
A_i = D^{-1}+\sigma^{-2}Z_i^{\!\top}W_i Z_i.
}
\tag{2}
\]

which is algebraically identical to the paper’s expression but only requires solving a \(q\times q\) system instead of an \(n_i\times n_i\) system.

In R

```{r}
# b_fun_small: optimized GMERT random-effects update (small q×q system)
b_fun_small <- function(G,        # number of clusters
                        Z,        # N x q random-effects design matrix
                        w,        # length N weights 
                        D,        # q x q covariance of random effects
                        sigma2,   # residual variance
                        idx,      # list of length G, indices per cluster
                        y_t,      # pseudo-response vector
                        fhat      # fitted fixed-part predictions
){
  b <- matrix(0, G, ncol(Z))      # store b_i estimates (G x q)

  for (g in seq_len(G)) {
    # --- extract cluster-specific data ---
    Zi <- Z[idx[[g]], , drop = FALSE]      # cluster design (n_i x q)
    y_res <- y_t[idx[[g]]] - fhat[idx[[g]]]# residuals (y_t - fhat)

    # --- compute small q×q system components ---
    # A_i = D^{-1} + (1/σ²) * Z_i^T W_i Z_i
    # rhs = (1/σ²) * Z_i^T W_i (y_t - fhat)
    A   <- solve(D) + (1 / sigma2) * crossprod(Zi, w[idx[[g]]] * Zi)
    rhs <- (1 / sigma2) * crossprod(Zi, w[idx[[g]]] * y_res)

    # --- solve for b_i via Cholesky (stable) ---
    R <- chol(A)
    b[g, ] <- backsolve(R, forwardsolve(t(R), rhs))
  }
  b
}

```


---

### 4. Trace of \(V_i^{-1}\)

Using (1) and the cyclic property of the trace, \(\operatorname{tr}(ABC)=\operatorname{tr}(BCA)=\operatorname{tr}(CAB)\) (with \(A = W_i^{1/2}Z_i\), \(B=A_i^{-1}\), \(C=Z_i^{\!\top}W_i^{1/2}\)), we have

\[
\operatorname{tr}(V_i^{-1})
= \operatorname{tr}\big(\sigma^{-2}I_{n_i}\big) - \operatorname{tr}\big(\sigma^{-4}W_i^{1/2}Z_i A_i^{-1} Z_i^{\!\top}W_i^{1/2}\big)
= \frac{n_i}{\sigma^2}
- \frac{1}{\sigma^4}\operatorname{tr}(A_i^{-1} Z_i^{\!\top}W_i Z_i).
\tag{3}
\]
Thus, the expensive \(n_i\times n_i\) trace can be replaced by a \(q\times q\) trace.

---

### 5. Small-System Expression for \(\sigma^2\)

Plug (3) into the paper’s \(\sigma^2\) update:
\[
\begin{aligned}
\hat\sigma^2
&=\frac{1}{N}\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \sigma^2\big(n_i-\sigma^2\,\operatorname{tr}(V_i^{-1})\big)
\Big] \\[4pt]
&=\frac{1}{N}\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \sigma^2\big(
n_i-\sigma^2\big(\tfrac{n_i}{\sigma^2}-\tfrac{1}{\sigma^4}\operatorname{tr}(A_i^{-1}Z_i^{\!\top}W_i Z_i)\big)
\big)
\Big] \\[4pt]
&=\frac{1}{N}\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \operatorname{tr}(A_i^{-1}Z_i^{\!\top}W_i Z_i)
\Big].
\end{aligned}
\]

Hence the compact small-system form:
\[
\boxed{
\hat\sigma^2
= \frac{1}{N}
\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \operatorname{tr}(A_i^{-1}Z_i^{\!\top}W_i Z_i)
\Big].
}
\tag{4}
\]

---

### 6. Small-System Expression for \(D\)

From the paper,
\[
\hat D = \frac{1}{n}\sum_i\big[b_i b_i^{\!\top} + (D - D(W^{1/2}Z_i)^{\!\top}V_i^{-1}W^{1/2}Z_i D)\big].
\]
Use (1) and \(U_i=W_i^{1/2}Z_i\):
\[
D - D U_i^{\!\top}V_i^{-1}U_i D
= D - D(\sigma^{-2}B_i - \sigma^{-4}B_iA_i^{-1}B_i)D,
\quad B_i=Z_i^{\!\top}W_i Z_i.
\]
Applying the identity
\[
D - D(\sigma^{-2}B_i)D
= (D^{-1}+\sigma^{-2}B_i)^{-1}=A_i^{-1},
\]
we obtain
\[
\boxed{
D - D(W^{1/2}Z_i)^{\!\top}V_i^{-1}W^{1/2}Z_i D
= A_i^{-1}.
}
\tag{5}
\]
Substituting (5) into the update gives the final efficient form:
\[
\boxed{
\hat D
= \frac{1}{n}\sum_i \big[\hat b_i\hat b_i^{\!\top} + A_i^{-1}\big].
}
\tag{6}
\]

---

## Summary of Small-System Formulas

All three updates can be computed using only \(q\times q\) matrix inverses, this reducing computational cost from \(O(n_i^3)\) to \(O(q^3)\) per group \(i\).
