---
title: "optimization"
author: "Paolo Colussi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("functions.R")
```
## Appendix – Derivation of the Small-System Formulas

One of the things that makes the GMERT algorithm computationally expensive is the need to invert the matrix \(\hat V_i\) of size \(n_i\times n_i\) at each iteration for each group \(i\) for the updates of the random effects \(\hat b_i\), the residual variance \(\hat\sigma^2\), and the random-effects covariance \(\hat D\). What can make this feasible is that the number of random effects \(q\) is typically much smaller than the number of observations per group \(n_i\). Using the Woodbury matrix identity, we can rewrite these updates in terms of inverses of \(q\times q\) matrices instead of \(n_i\times n_i\) matrices, which is computationally much cheaper. This appendix derives these small-system formulas.

### 1. Original GMERT Updates

First, we restate the relevant definitions from the GMERT paper. 
At iteration \(m\), the paper defines the random-effects, residual, and covariance updates as:

\[
\begin{aligned}
\hat b_i^{(m)} &=
\hat D_{(m-1)} (W_i^{1/2(M)} Z_i)^{\!\top}
\hat V_i^{-1(m-1)}
W_i^{1/2(M)}\big(\tilde y_i^{(M)} - \hat f_{(m)}(X_i)\big), \\[4pt]
\hat V_i^{(m-1)} &=
W_i^{1/2(M)} Z_i \hat D_{(m-1)} Z_i^{\!\top} W_i^{1/2(M)}
+ \hat\sigma_{(m-1)}^2 I_{n_i}, \\[6pt]
\hat\sigma^2_{(m)} &=
\frac{1}{N} \sum_{i=1}^{n}
\Big\{
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \hat\sigma^2_{(m-1)}\big[n_i - \hat\sigma^2_{(m-1)} \operatorname{tr}(\hat V_i^{-1(m-1)})\big]
\Big\}, \\[6pt]
\hat D_{(m)} &=
\frac{1}{n}
\sum_{i=1}^n
\Big\{
\hat b_i^{(m)} \hat b_i^{(m)\top}
+
\big[
\hat D_{(m-1)} -
\hat D_{(m-1)}(W_i^{1/2}Z_i)^{\!\top}
\hat V_i^{-1(m-1)} W_i^{1/2}Z_i \hat D_{(m-1)}
\big]
\Big\},
\end{aligned}
\]
with
\[
\hat\epsilon_i^{(m)} = W_i^{1/2}\big(\tilde y_i - \hat f_i - Z_i \hat b_i^{(m)}\big).
\]

All inverses are over the \(n_i\times n_i\) matrices \(\hat V_i\), which dominates computational cost.

---

### 2. Woodbury Identity

To derive small-system expressions, we use the Woodbury matrix identity:
For matrices \(A\in\mathbb{R}^{n_i\times n_i}\), \(U\in\mathbb{R}^{n_i\times q}\),
and \(C\in\mathbb{R}^{q\times q}\),
\[
(A + UCU^{\!\top})^{-1}
= A^{-1} - A^{-1}U\,(C^{-1}+U^{\!\top}A^{-1}U)^{-1}U^{\!\top}A^{-1}.
\]

Set \(A=\sigma^2 I_{n_i}\), \(U=W_i^{1/2}Z_i\), \(C=D\).
Then
\[
\boxed{
V_i^{-1}
= \sigma^{-2}I_{n_i}
- \sigma^{-4}W_i^{1/2}Z_i
\big(D^{-1}+\sigma^{-2}Z_i^{\!\top}W_i Z_i\big)^{-1}
Z_i^{\!\top}W_i^{1/2}.
}
\tag{1}
\]

Define
\[
A_i = D^{-1} + \sigma^{-2} Z_i^{\!\top} W_i Z_i
\quad\text{(\(q\times q\) matrix).}
\]

In R we can implement \(A_i^{-1}\) as follows:

```{r}
# Ainv_fun: precomputes A_i^{-1} (and optionally Z_i^T W_i Z_i) for all clusters
Ainv_fun <- function(G,        # G      : number of clusters
                     Z,        # Z      : N x q random-effects design
                     w,        # W      : list of weigths 
                     D,        # D      : q x q covariance of random effects (current)
                     sigma2,   # sigma2 : residual variance (current)
                     idx      # idx    : list length G with row indices for each cluster
){
  Ainv_list <- vector("list", G)

  Dinv <- solve(D)                                 # D^{-1} (q x q) once

  for (g in seq_len(G)) {
    Zi  <- Z[idx[[g]], , drop = FALSE]             # n_i x q
    Wi  <- diag(w[idx[[g]]])                      
    ZtWZ <- t(Zi) %*% Wi %*% Zi                     # Z_i^T W_i Z_i (q x q)

    A <- Dinv + as.numeric((1 / sigma2)) * ZtWZ                # A_i
    Ainv <- solve(A)                              # A_i^{-1} 

    Ainv_list[[g]] <- Ainv
  }

  Ainv_list
}

```

We calculate directly \(A_i^{-1}\) for each group \(i\) since it is needed in all three updates, and we don't have to solve the same system three times.

---

### 3. Small-System Expression for \(b_i\)

Substituting (1) into the paper’s definition of \(\hat b_i\)
and simplifying with \(W_i^{1/2}Z_i=U_i\)
gives

\[
\begin{aligned}
\hat b_i
&= D (W_i^{1/2}Z_i)^{\!\top}
\big[\sigma^{-2}I - \sigma^{-4}W_i^{1/2}Z_iA_i^{-1}Z_i^{\!\top}W_i^{1/2}\big]
W_i^{1/2}(\tilde y_i-\hat f_i) \\[3pt]
&= \big(D^{-1}+\sigma^{-2}Z_i^{\!\top}W_i Z_i\big)^{-1}
      \big(\sigma^{-2}Z_i^{\!\top}W_i(\tilde y_i-\hat f_i)\big).
\end{aligned}
\]

Hence,
\[
\boxed{
\hat b_i =
A_i^{-1}\,\sigma^{-2}Z_i^{\!\top}W_i(\tilde y_i-\hat f_i),
\qquad
A_i = D^{-1}+\sigma^{-2}Z_i^{\!\top}W_i Z_i.
}
\tag{2}
\]

which is algebraically identical to the paper’s expression but only requires solving a \(q\times q\) system instead of an \(n_i\times n_i\) system.

In R

```{r}
# b_fun_Ainv: updates random effects b_i using precomputed A_i^{-1}
b_fun_small <- function(G,        # G      : number of clusters
                       Z,        # Z      : N x q random-effects design
                       w,        # W      : list of weigths 
                       idx,      # idx    : list length G with row indices for each cluster
                       y_t,      # y_t    : pseudo-response vector (length N)
                       fhat,     # fhat   : fitted fixed-part prediction (length N)
                       Ainv,     # Ainv   : list length G; each Ainv[[g]] = A_i^{-1} (q x q)
                       sigma2    # sigma2 : residual variance (current)
){
  q <- ncol(Z)
  b <- matrix(0, G, q)

  for (g in seq_len(G)) {
    Zi   <- Z[idx[[g]], , drop = FALSE]           # n_i x q
    Wi   <- diag(w[idx[[g]]])                     # weights vector (length n_i)
    resid<- y_t[idx[[g]]] - fhat[idx[[g]]]        # (y_t - fhat) on cluster i

    rhs  <- as.numeric((1 / sigma2)) * t(Zi) %*% (Wi %*% resid)  # right-hand side
    b[g, ] <- as.vector(Ainv[[g]] %*% rhs)           # b_i = A_i^{-1} * rhs
  }

  b
}

```


---

### 4. Trace of \(V_i^{-1}\)

Using (1) and the cyclic property of the trace, \(\operatorname{tr}(ABC)=\operatorname{tr}(BCA)=\operatorname{tr}(CAB)\) (with \(A = W_i^{1/2}Z_i\), \(B=A_i^{-1}\), \(C=Z_i^{\!\top}W_i^{1/2}\)), we have

\[
\operatorname{tr}(V_i^{-1})
= \operatorname{tr}\big(\sigma^{-2}I_{n_i}\big) - \operatorname{tr}\big(\sigma^{-4}W_i^{1/2}Z_i A_i^{-1} Z_i^{\!\top}W_i^{1/2}\big)
= \frac{n_i}{\sigma^2}
- \frac{1}{\sigma^4}\operatorname{tr}(A_i^{-1} Z_i^{\!\top}W_i Z_i).
\tag{3}
\]
Thus, the expensive \(n_i\times n_i\) trace can be replaced by a \(q\times q\) trace.

---

### 5. Small-System Expression for \(\sigma^2\)

Plug (3) into the paper’s \(\sigma^2\) update:
\[
\begin{aligned}
\hat\sigma^2
&=\frac{1}{N}\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \sigma^2\big(n_i-\sigma^2\,\operatorname{tr}(V_i^{-1})\big)
\Big] \\[4pt]
&=\frac{1}{N}\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \sigma^2\big(
n_i-\sigma^2\big(\tfrac{n_i}{\sigma^2}-\tfrac{1}{\sigma^4}\operatorname{tr}(A_i^{-1}Z_i^{\!\top}W_i Z_i)\big)
\big)
\Big] \\[4pt]
&=\frac{1}{N}\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \operatorname{tr}(A_i^{-1}Z_i^{\!\top}W_i Z_i)
\Big].
\end{aligned}
\]

Hence the compact small-system form:
\[
\boxed{
\hat\sigma^2
= \frac{1}{N}
\sum_i
\Big[
\hat\epsilon_i^{\!\top}\hat\epsilon_i
+ \operatorname{tr}(A_i^{-1}Z_i^{\!\top}W_i Z_i)
\Big].
}
\tag{4}
\]

In R

```{r}
# sigma_fun_Ainv: updates residual variance sigma2 using precomputed A_i^{-1}
sigma_fun_small <- function(N,        # N      : total number of rows
                           G,        # G      : number of clusters
                           idx,      # idx    : list length G with row indices for each cluster
                           Z,        # Z      : N x q random-effects design
                           w,        # W      : list of weigths
                           y_t,      # y_t    : pseudo-response vector
                           fhat,     # fhat   : fitted fixed-part prediction
                           b,        # b      : matrix G x q with current random effects
                           Ainv,     # Ainv   : list length G; each Ainv[[g]] = A_i^{-1}
                           sigma2   # sigma2 : residual variance (current)
){
  rss_total <- 0

  for (g in seq_len(G)) {
    Zi  <- Z[idx[[g]], , drop = FALSE]             # n_i x q
    wi  <- diag(w[idx[[g]]])                             # weights matrix 
    ni  <- length(wi)

    # residuals eps_i
    eps_i <- diag(sqrt(w[idx[[g]]])) %*% (y_t[idx[[g]]] - fhat[idx[[g]]] - as.vector(Zi %*% b[g, ]))

    # tr(V_i^{-1})
    trVi_inv <- (ni / sigma2) - as.numeric((1 / sigma2^2)) * sum(diag(Ainv[[g]] %*% t(Zi) %*% wi %*% Zi))

    # accumulate rss
    rss_total <- rss_total + t(eps_i) %*% eps_i + sigma2 * (ni - sigma2 * trVi_inv)
  }

  rss_total / N
}

```


---

### 6. Small-System Expression for \(D\)

From the paper,
\[
\hat D = \frac{1}{n}\sum_i\big[b_i b_i^{\!\top} + (D - D(W^{1/2}Z_i)^{\!\top}V_i^{-1}W^{1/2}Z_i D)\big].
\]
Use (1) and \(U_i=W_i^{1/2}Z_i\):
\[
D - D U_i^{\!\top}V_i^{-1}U_i D
= D - D(\sigma^{-2}B_i - \sigma^{-4}B_iA_i^{-1}B_i)D,
\quad B_i=Z_i^{\!\top}W_i Z_i.
\]
Applying the identity
\[
D - D(\sigma^{-2}B_i)D
= (D^{-1}+\sigma^{-2}B_i)^{-1}=A_i^{-1},
\]
we obtain
\[
\boxed{
D - D(W^{1/2}Z_i)^{\!\top}V_i^{-1}W^{1/2}Z_i D
= A_i^{-1}.
}
\tag{5}
\]
Substituting (5) into the update gives the final efficient form:
\[
\boxed{
\hat D
= \frac{1}{n}\sum_i \big[\hat b_i\hat b_i^{\!\top} + A_i^{-1}\big].
}
\tag{6}
\]

In R

```{r}
# D_fun_Ainv: updates random-effects covariance D using precomputed A_i^{-1}
D_fun_small <- function(G,        # G      : number of clusters
                       b,        # b      : matrix G x q with current random effects
                       Ainv      # Ainv   : list length G; each Ainv[[g]] = A_i^{-1}
){
  q <- ncol(b)
  D_new <- matrix(0, q, q)

  for (g in seq_len(G)) {
    D_new <- D_new + tcrossprod(b[g, ]) + Ainv[[g]]   # b_i b_i^T + A_i^{-1}
  }

  D_new / G
}

```


---

### Summary of Small-System Formulas

All three updates can be computed using only \(q\times q\) matrix inverses, this reducing computational cost from \(O(n_i^3)\) to \(O(q^3)\) per group \(i\). 

We can now implement on R a new version of the GMERT updates using these small-system formulas. It is basically the same as the version in the gmert_algo.Rmd, except that we replace the b_fun, sigma_fun, and D_fun functions with the small-system versions derived here, and we add a call to Ainv_fun to precompute the \(A_i^{-1}\) matrices at each iteration. So we first of all have to call the function to compute the GLL.

```{r}
fit_gmert_small    <- function(df,         # df: data.frame with columns
                                           #   id   : cluster identifier (factor or integer)
                                           #   y    : numeric response
                                           #   x1,x2,x3 : numeric covariates
                                           #   leaf : true generating leaf (optional, for diagnostics)
                         max_iter_inn = 1000,  # maximum number of EM iterations (inner loop)
                         max_iter_out = 1000,  # maximum number of PQL iterations (outer loop)
                         tol = 1e-6,           # convergence tolerance for both loops (Aitken or relative diff)
                         cp = 0.0,             # rpart complexity parameter (pruning threshold)
                         minsplit = 50,        # minimum number of obs required to attempt a split
                         minbucket = 20,       # minimum number of obs in any terminal node
                         maxdepth = 5,         # maximum tree depth
                         xval = 10             # number of cross-validation folds in rpart
                         ) {

  # --- Basic setup ---
  N <- nrow(df)                                 # total number of observations
  G <- length(unique(df$id))                    # number of clusters
  idx_by_cluster <- split(seq_len(N), df$id)    # list: row indices grouped by cluster

  y <- df$y                                     # response vector
  Z <- cbind(1, df$x1)                          # random-effects design: intercept + slope on x1
  q <- ncol(Z)                                  # number of random effects (q = 2)

  # --- Initialization (Step 0) ---
  M <- 0L                                       # outer-loop counter
  mu <- ifelse(y == 1, 0.75, 0.25)              # initial conditional means
  y_t <- log(mu / (1 - mu)) + (y - mu) / (mu * (1 - mu))  # initial pseudo-response (PQL linearization)
  w <- mu * (1 - mu)                            # initial working weights
  sigma2 <- 1                                   # initial residual variance
  D <- diag(2)                                  # initial random-effects covariance (identity)
  b <- matrix(0, G, q)                          # initialize cluster random effects
  gll <- c(0, 0)                                # GLL storage (2 slots for Aitken acceleration)
  eta_old <- rep(0, N)                          # previous eta for outer-loop convergence check
  converged_in <- c()                           # inner-loop convergence flags (per outer iteration)
  converged_out <- FALSE                        # outer-loop convergence flag

  # --- Outer loop (PQL updates) ---
  repeat {
    m = 0                                       # reset inner-loop counter

    # --- Inner loop (EM-like iterations) ---
    repeat {
      m <- m + 1L

      # (1.i) Partial E-step: compute adjusted pseudo-response y_tilde* = y_tilde - Z b
      zb <- numeric(N)                          # cluster-specific random contributions
      for (g in seq_len(G)) {
        idx <- idx_by_cluster[[g]]              # indices for cluster g
        zb[idx] <- Z[idx, , drop = FALSE] %*% b[g, ]  # Z_i b_i
      }
      y_star <- y_t - zb                        # adjusted response for tree fit

      # (1.ii) M-step: fit regression tree for fixed effects f(X)
      ctrl <- rpart.control(cp = cp, minsplit = minsplit, xval = xval,
                            minbucket = minbucket, maxdepth = maxdepth)
      Xdf <- data.frame(x1 = df$x1, x2 = df$x2, x3 = df$x3)
      tree <- rpart(y_star ~ x1 + x2 + x3,
                    data = cbind(y_star = y_star, Xdf),
                    weights = w, method = "anova", control = ctrl)
      fhat <- as.numeric(predict(tree, newdata = Xdf))

      # (1.iii) Update random effects b_i
      Ainv <- Ainv_fun(Z = Z, D = D, sigma2 = sigma2, G = G, idx = idx_by_cluster, w = w)  # build V_i per cluster
      b <- b_fun_small(G = G, Z = Z, idx = idx_by_cluster,
                 y_t = y_t, fhat = fhat, w = w, sigma2 = sigma2, Ainv = Ainv)                                   # update b_i estimates

      # (2.i) Update sigma^2 (residual variance)
      sigma2 <- sigma_fun_small(N = N, G = G, idx = idx_by_cluster,
                          b = b, y = y_t, Z = Z, 
                          Ainv = Ainv, sigma2 = sigma2, fhat = fhat, w = w)

      # (2.ii) Update D (random-effects covariance)
      D <- D_fun_small(G = G, b = b, Ainv = Ainv)

      # --- Inner-loop convergence check (GLL stabilization) ---
      gll[m + 2] <- gll_fun(D = D, b = b, idx = idx_by_cluster,
                            Z = Z, y = y_t, fhat = fhat, s2 = sigma2, w = w)
      if (m > 1L) {
        rel <- abs(gll[m + 2] - gll[m + 1]) / (abs(gll[m + 1]) + 1e-12)
        if (rel < tol) { n_iter <- m; converged_in_t <- TRUE; break }
      }

      if (m >= max_iter_inn) {                  # max-iteration guard
        n_iter <- m
        converged_in_t <- FALSE
        # message(sprintf("WARNING: the EM algorithm did not converge in %d iterations.", max_iter_inn))
        break
      }
    }

    # --- Outer-loop update (PQL step) ---
    converged_in <- c(converged_in, converged_in_t)
    zb <- numeric(N)
    for (g in seq_len(G)) {
      idx <- idx_by_cluster[[g]]
      zb[idx] <- Z[idx, , drop = FALSE] %*% b[g, ]
    }
    eta <- fhat + zb                            # recompute linear predictor

    # (Outer stopping rule – paper style)
    d_eta <- sqrt(mean((eta - eta_old)^2))      # RMS change of eta
    if (d_eta < tol) {
      converged_out <- TRUE
      break
    }

    # Update working quantities for next outer iteration
    eta_old <- eta
    mu <- exp(eta) / (1 + exp(eta))             # updated conditional means
    y_t <- log(mu / (1 - mu)) + (y - mu) / (mu * (1 - mu))  # new pseudo-response
    w <- mu * (1 - mu)                          # new working weights

    M <- M + 1L
    if (M >= max_iter_out) {                    # guard against outer non-convergence
      converged_out <- FALSE
      message(sprintf("WARNING: the PQL algorithm did not converge in %d iterations.", max_iter_out))
      break
    }
  }

  # --- Return fitted components ---
  list(
    tree = tree,                   # fitted rpart tree (fixed-effects function f(X))
    b = b,                         # estimated random effects (G x q)
    D = D,                         # estimated random-effects covariance
    sigma2 = sigma2,               # estimated residual variance
    mu = mu,                       # conditional means
    converged_in = converged_in,   # convergence flags for inner loops
    converged_out = converged_out, # convergence flag for outer loop
    n_iter = n_iter,               # iterations performed (inner loop)
    train_ids = df$id,             # cluster identifiers in training set
    gll = gll,                     # GLL trace (for diagnostics)
    tol = tol                      # convergence tolerance used
  )
}

```

## Testing

```{r}
#  Simulate clustered binary data (30 clusters, 60 observations each)
df <- sim_data_gmert(G = 30, n_i = 60)

#  Fit the GMERT model using the full dataset
system.time(fit_gmert(df, max_iter_inn = 1000, max_iter_out = 10, tol = 1e-6))

system.time(fit_gmert_small(df, max_iter_inn = 1000, max_iter_out = 10, tol = 1e-6))

```



